---
title: "Friedman Chi-Square test using R package rstatix"
---

## R package version

R 4.3.1\
rstatix 0.7.2

## Data used

Simulated dataset of 10 subjects(blocks) with continuous endpoints are generated for single-drug repeated measurements to check whether any significance exists between the responses(y) at different time points(4 time points simulated)(groups). The p-value will indicate whether differences in response for different time points are significant.

## Data source

```{r eval=TRUE}
library(dplyr)
library(rstatix) 
```

```{r eval=TRUE}
set.seed(123)

one_way_repeat <- data.frame(subject = rep(1:10, each=4),
                          timepoint = rep(c(1, 2, 3, 4), times=10),
                          response =round(runif(n = 40, 10, 50)))

head(one_way_repeat)
```

## Overview

The friedman.test() function from the R package 'rstatix' computes the following:

-   Friedman chi-squared statistic

-   its degrees of freedom and

-   p-value for checking whether the mean response is the same for all the time points.

## Handling missing Values

When the data contain NAs, the function Defaults to getOption("na.action") which discards the row with NAs.

## Example Code for Friedman Chi-square test

```{r eval=TRUE}
friedman.test(y=one_way_repeat$response, groups=one_way_repeat$timepoint, blocks=one_way_repeat$subject)
```

where:

-   y - a vector of response values.

-   groups - a vector of values indicating the "group" an observation belongs in.

-   blocks - a vector of values indicating the "blocking" variable.

## References

[R `friedman_test()` documentation](https://search.r-project.org/CRAN/refmans/rstatix/html/friedman_test.html)
