---
title: "git training graph code"
---


#@christina: take this for the Week 1 Day 1 graph: we are creating a publication for PLOS One
#Camille has 2 journals in mind for her submissions,  she reads the specifciations for each and then starts prorgramming.
#PLOS ONE journal requires a font of: Arial, Times, or Symbol font only in 8-12 point
#Journal of Statistical Theory and Practice: reccommends using Helvetica or Arial (sans serif fonts).


#read in data: survival data and create variables for labels

```{r}
#| echo: false
#| include: false
library(tidyverse)
library(broom)
library(knitr)
library(haven)

surv<-read_sas("../../data/whas500.sas7bdat")%>%
  mutate(LENFOLY = round(LENFOL/365.25, 2), ## change follow-up days to years for better visualization
         AFB_C = ifelse(AFB==1, "Active", "Non-Active"),
         GENDER_C = ifelse(GENDER==1, "Male", "Female"),
         CVD_C = ifelse(CVD==1, "Yes", "No")) 
```

#Create a graph of drug response by Presence of absence of cardiovascular disease.
# This is submitted to PLOS ONE journal on Tuesday.
```{r}
  ggplot(data = surv, mapping = aes(x =SYSBP , y =DIASBP )) + 
  geom_point() + 
  facet_grid(rows = vars(CVD_C), cols = vars(GENDER_C))+

  labs(title="Diastolic and Systolic Blood Pressure by Gender and Cardiovascular Disease Status",
       x="Systolic Blood Pressure",
       y="Diastolic Blood Pressure",
       caption="Data source: Survival 500 data")+
  geom_smooth(mapping=aes(group=NA, x=SYSBP, y=DIASBP))+
  theme(panel.spacing = unit(0.5,"cm",data=NULL),
        strip.text.x=element_text(size=14, color="red",face="bold.italic"),
        strip.text.y=element_text(size=7, color="blue",face="bold")
  )


```

# on Wednesday Camille discusses the results with a colleague, they think the results may not be accepted by journal
# as not novel. Instead they suggest Camille explores the Status of disease being active or non-active.
# Camille continues to explore the data and makes changes substantial changes to her figure which she
# thinks looks much prettier !!!  She saves the program and leaves for the weekend
```{r}
  ggplot(data = surv, mapping = aes(x =DIASBP , y =SYSBP )) + 
  geom_point() + 
  facet_grid(rows = vars(AFB_C), cols = vars(GENDER_C))+

  labs(title="Systolic and Diasolic Blood Pressure by Gender and Disease Status",
       y="Systolic Blood Pressure",
       x="Diastolic Blood Pressure",
       caption="Data source: Survival 500 data")+
  geom_smooth(mapping=aes(group=NA,  x=DIASBP,y=SYSBP))+
  theme(panel.spacing = unit(0.5,"cm",data=NULL),
        strip.text.x=element_text(size=12, color="blue",face="bold.italic"),
        strip.text.y=element_text(size=12, color="blue",face="bold"),
        plot.title=element_text(face="italic", color="Blue"),
        panel.grid.minor=element_line(linetype="solid",color="black"),
        plot.background=element_rect(fill="#80ff00",color="black")
  )

```

##  Week 2, Camille gets an email from the publication stressing that she hasn't following the guidance here:
#https://journals.plos.org/plosone/s/figures
# All figures much be produced using font sizes between 8 and 12.
#if she could find her previous code it would be a simple update, however, the code has changed so much
#it's goign to take her ages to recreate it. 

#PLOS ONE waives all formatting requirements until your manuscript has received a provisional Editorial Accept decision.
#In order to proceed to publication, your figures must meet the requirements on this page. The more closely your figures #adhere to these specifications, the more quickly your manuscript can be published once accepted.
#The guidance states to use Arial, Times or Symbol font only in 8-12 point.

# her application is rejected.

