---
title: "R/SAS Chi-Squared and Fisher's Exact Comparision"
---

# Chi-Squared Test

Chi-Squared test is a hypothesis test for independent contingency tables, dependent on rows and column totals. The test assumes:

-   observations are independent of each other

-   all values are 1 or more and at least 80% of the cells are greater than 5.

-   data should be categorical

The Chi-Squared statistic is found by:

$$
\chi^2=\frac{\sum(O-E)^2}{E}
$$

Where O is the observed and E is the expected.\
For an r x c table (where r is the number of rows and c the number of columns), the Chi-squared distribution's degrees of freedom is (r-1)\*(c-1). The resultant statistic with correct degrees of freedom follows this distribution when its expected values are aligned with the assumptions of the test, under the null hypothesis. The resultant p value informs the magnitude of disagreement with the null hypothesis and not the magnitude of association

For this example we will use data about cough symptoms and history of bronchitis.

```{r}

bronch <- matrix(c(26, 247, 44, 1002), ncol = 2)
row.names(bronch) <- c("cough", "no cough")
colnames(bronch) <- c("bronchitis", "no bronchitis")
```

To a chi-squared test in R you will use the following code.

```{r}
chisq.test(bronch)
```

To run a chi-squared test in SAS you used the following code.

```{SAS}
#| eval: false
proc freq data=proj1.bronchitis;
tables Cough*Bronchitis / chisq;
run;
```

The result in the "Chi-Square" section of the results table in SAS will not match R, in this case it will be 12.1804 with a p-value of 0.0005. This is because by default R does a Yate's continuity adjustment. To change this set `correct` to false.

```{r}
chisq.test(bronch, correct = FALSE)
```

Alternatively, SAS also produces the correct chi-square value by default. It is the "Continuity Adj. Chi-Square" value in the results table.

# Fisher's Exact Test

Comparison between the Fisher's Exact Test in both R and SAS shows that the two
software match on the p-value and confidence intervals.  The odd ratio does not match.
The reason the odds ratio does not match is because R uses an "exact" odds ratio based 
on the hypergeomtric distribution, while SAS uses a standard AD/BC odds ratio.  Note
that R always uses an "exact" Fisher test.  Therefore, when trying to match SAS, you
must use the "exact" statement on the PROC FREQ.