---
title: "R vs SAS vs Python Survey Summary Statistics"
bibliography: survey-stats-summary.bib
---

This document will compare the survey summary statistics functionality in SAS (available through SAS/STAT), R (available from the [`{survey}`](%5B%60%7Bsurvey%7D%60%5D(https://r-survey.r-forge.r-project.org/survey/html/api.html)) package), and Python (available from the [`samplics`](https://samplics-org.github.io/samplics/) package), highlighting differences in methods and results. Only the default Taylor series linearisation method for calculating variances is used in all languages. A more detailed comparison between R and SAS for specific methods and use-cases is available in [@2017_YRBS], [@so2020modelling], or [@adamico_2009]. For a general guide to survey statistics, which has companion guides for both R and SAS, see [@Lohr_2022].

# Result Comparison

The following table shows different survey summary statistics, the capabilities of each language, and whether or not the results match. Each analysis also includes calculating the standard error and confidence intervals.

| Analysis        | Supported in R                                                                               | Supported in SAS                                                                                | Supported in Python                                             | Results Match\*  | Notes                                                                                                                                          |
|------------|------------|------------|------------|------------|------------|
| Mean            | [Yes](../R/survey-stats-summary.html#Mean)                                                   | [Yes](../SAS/survey-stats-summary.html#Mean)                                                    | [Yes](../Python/survey-stats-summary.html#Mean)                 | Yes              | Must specify degrees of freedom in R for confidence limits                                                                                     |
| Total           | [Yes](../R/survey-stats-summary.html#Total)                                                  | [Yes](../SAS/survey-stats-summary.html#Total)                                                   | [Yes](../Python/survey-stats-summary.html#Total)                | Yes              | Must specify degrees of freedom in R for confidence limits                                                                                     |
| Ratios          | [Yes](../R/survey-stats-summary.html#Ratios)                                                 | [Yes](../SAS/survey-stats-summary.html#Ratios)                                                  | [Yes](../Python/survey-stats-summary.html#Ratios)               | Yes              | Must specify degrees of freedom in R for confidence limits                                                                                     |
| Proportions     | [Yes](../R/survey-stats-summary.html#Proportions)                                            | [Yes](../SAS/survey-stats-summary.html##Proportions)                                            | [Yes](../Python/survey-stats-summary.html#Proportions)          | Yes\*\*          | In Python, the confidence limits of proportions only match to 1 or 2 s.f. This is due to a different method being used, which is undocumented. |
| Quantiles       | [Yes](../R/survey-stats-summary.html#Quantiles)                                              | [Yes](../SAS/survey-stats-summary.html#Quantiles)                                               | No                                                              | [No](#Quantiles) | Different methods for calculating quantiles                                                                                                    |
| Domain Analysis | [Yes](../R/survey-stats-summary.html#Summary%20Statistics%20on%20Complex%20Survey%20Designs) | [Yes](../SAS/survey-stats-summary.html##Summary%20Statistics%20on%20Complex%20Survey%20Designs) | [Yes](../Python/survey-stats-summary.html#Domain%20Estimations) | Yes              |                                                                                                                                                |
| Design Effect   | [Yes](../R/survey-stats-summary.html#Summary%20Statistics%20on%20Complex%20Survey%20Designs) | [Yes](../SAS/survey-stats-summary.html##Summary%20Statistics%20on%20Complex%20Survey%20Designs) | No                                                              | Yes              | Set `deff="replace"` in R to match SAS exactly                                                                                                 |

*\*Results match where feature is available*

*\*\*For confidence limits of proportions near 0 and 1, `survey::svyciprop` can be more accurate than `confint` in R, but does not match other software.*

For the full R, SAS, and Python code and results used for this comparison, see below:

::: {.callout-note collapse="true" appearance="minimal" title="Show Code"}
## R

```{r}
#| message: false
#| warning: false 
library(survey)

data("nhanes")

nhanes_design <- svydesign(
  data = nhanes,
  id = ~SDMVPSU, # Specify the PSU/cluster column
  strata = ~SDMVSTRA,  # The stratification column
  weights = ~WTMEC2YR,  # The weighting column
  nest = TRUE  # Allows for PSUs with the same name nested within different strata
)

# Mean of HI_CHOL
hi_chol_mean <- svymean(~HI_CHOL, nhanes_design, na.rm=TRUE)

# Sum of HI_CHOL
hi_chol_sum <- svytotal(~HI_CHOL, nhanes_design, na.rm=TRUE)

# Ratio of HI_CHOL / RIAGENDR
hi_chol_ratio <- svyratio(
  numerator = ~HI_CHOL,
  denominator = ~RIAGENDR,
  nhanes_design,
  na.rm=TRUE,
  ci=TRUE,
  se=TRUE,
  separate=FALSE
)

# Proportion of different AGECAT values
agecat_props <- svymean(~agecat, nhanes_design, na.rm=TRUE)

# Quantiles of HI_CHOL
hi_chol_quart <- svyquantile(
  ~HI_CHOL,
  nhanes_design,
  quantiles=c(0.025, 0.5, 0.975),
  na.rm=TRUE,
  ci=TRUE
)

# Domain analysis of mean of HI_CHOL by race, with design effect
hi_chol_mean_by_race <- svyby(~HI_CHOL, ~race, nhanes_design, svymean, na.rm=TRUE, deff="replace")

print(list(
  "Mean of HI_CHOL" = coef(hi_chol_mean),
  "SE of Mean HI_CHOL" = SE(hi_chol_mean),
  "CL of Mean HI_CHOL" = confint(hi_chol_mean, df=degf(nhanes_design)),
  "Sum of HI_CHOL" = coef(hi_chol_sum),
  "SE of Sum HI_CHOL" = SE(hi_chol_sum),
  "CL of Sum HI_CHOL" = confint(hi_chol_sum, df=degf(nhanes_design)),
  "Ratio of HI_CHOL / RIAGENDR" = coef(hi_chol_ratio),
  "SE of Ratio HI_CHOL / RIAGENDR" = SE(hi_chol_ratio),
  "CL of Ratio HI_CHOL / RIAGENDR" = confint(hi_chol_ratio, df=degf(nhanes_design)),
  "Proportion of AGECAT" = coef(agecat_props),
  "SE of Proportion AGECAT" = SE(agecat_props),
  "CL of Proportion AGECAT" = confint(agecat_props, df=degf(nhanes_design)),
  "Quantiles of HI_CHOL" = coef(hi_chol_quart),
  "SE of Quantiles HI_CHOL" = SE(hi_chol_quart),
  "CL of Quantiles HI_CHOL" = confint(hi_chol_quart, df=degf(nhanes_design)),
  "Mean of HI_CHOL by race" = coef(hi_chol_mean_by_race),
  "SE of HI_CHOL by race" = SE(hi_chol_mean_by_race),
  "CL of HI_CHOL by race" = confint(hi_chol_mean_by_race, df=degf(nhanes_design)),
  "Design Effect of HI_CHOL by race" = hi_chol_mean_by_race$DEff.HI_CHOL
))
```

## SAS

``` default
* Mean, sum quantile of HI_CHOL;
proc surveymeans data=nhanes mean sum clm quantile=(0.025 0.5 0.975);
    cluster SDMVPSU;
    strata SDMVSTRA;
    weight WTMEC2YR;
    var HI_CHOL;
run;

* Ratio of HI_CHOL / RIAGENDR;
proc surveymeans data=nhanes;
    cluster SDMVPSU;
    strata SDMVSTRA;
    weight WTMEC2YR;
    ratio HI_CHOL / RIAGENDR;
run;

* Proportions of agecat;
proc surveyfreq data=nhanes;
    cluster SDMVPSU;
    strata SDMVSTRA;
    weight WTMEC2YR;
    table agecat / cl;
run;

* Mean and DEFF of HI_CHOL by race;
proc surveymeans data=nhanes mean deff;
    cluster SDMVPSU;
    strata SDMVSTRA;
    weight WTMEC2YR;
    domain race;
    var HI_CHOL;
run;
```

```         

                                                 The SURVEYMEANS Procedure

                                                        Data Summary

                                            Number of Strata                  15
                                            Number of Clusters                31
                                            Number of Observations          8591
                                            Sum of Weights             276536446


                                                         Statistics

                                Std Error                                                Std Error
 Variable            Mean         of Mean       95% CL for Mean                Sum          of Sum        95% CL for Sum
 --------------------------------------------------------------------------------------------------------------------------
 HI_CHOL         0.112143        0.005446    0.10059829 0.12368762        28635245         2020711    24351529.8 32918960.7
 --------------------------------------------------------------------------------------------------------------------------


                                                         Quantiles

                                                                          Std
                     Variable       Percentile       Estimate           Error    95% Confidence Limits
                     ---------------------------------------------------------------------------------
                     HI_CHOL          2.5                   0        0.024281    -0.0514730 0.05147298
                                       50 Median            0        0.024281    -0.0514730 0.05147298
                                     97.5            0.777070        0.024281     0.7255973 0.82854324
                     ---------------------------------------------------------------------------------

                                                 The SURVEYMEANS Procedure

                                                        Data Summary

                                            Number of Strata                  15
                                            Number of Clusters                31
                                            Number of Observations          8591
                                            Sum of Weights             276536446


                                                        Statistics

                                                                    Std Error
                     Variable               N            Mean         of Mean       95% CL for Mean
                     ---------------------------------------------------------------------------------
                     HI_CHOL             7846        0.112143        0.005446    0.10059829 0.12368762
                     RIAGENDR            8591        1.512019        0.005302    1.50077977 1.52325807
                     ---------------------------------------------------------------------------------


                                                       Ratio Analysis

                                                                              Std
               Numerator Denominator            N           Ratio           Error        95% CL for Ratio
               ----------------------------------------------------------------------------------------------
               HI_CHOL   RIAGENDR            7846        0.074222        0.003715    0.06634722    0.08209696
               ----------------------------------------------------------------------------------------------

                                                  The SURVEYFREQ Procedure

                                                        Data Summary

                                            Number of Strata                  15
                                            Number of Clusters                31
                                            Number of Observations          8591
                                            Sum of Weights             276536446


                                                      Table of agecat

                                        Weighted    Std Err of                Std Err of    95% Confidence Limits
          agecat         Frequency     Frequency      Wgt Freq     Percent       Percent         for Percent
          -------------------------------------------------------------------------------------------------------
          (0,19]              2532      57450307       3043819     20.7749        0.6130     19.4755      22.0744
          (19,39]             2033      81137975       3692818     29.3408        0.9561     27.3140      31.3676
          (39,59]             2021      83870623       4853936     30.3290        0.4519     29.3709      31.2870
          (59,Inf]            2005      54077541       4284296     19.5553        0.8093     17.8398      21.2709

          Total               8591     276536446      13935730    100.0000                                       
          -------------------------------------------------------------------------------------------------------

                                                 The SURVEYMEANS Procedure

                                                        Data Summary

                                            Number of Strata                  15
                                            Number of Clusters                31
                                            Number of Observations          8591
                                            Sum of Weights             276536446


                                                         Statistics

                                                                 Std Error          Design
                                  Variable            Mean         of Mean          Effect
                                  --------------------------------------------------------
                                  HI_CHOL         0.112143        0.005446        2.336725
                                  --------------------------------------------------------

                                                 The SURVEYMEANS Procedure

                                                 Statistics for race Domains

                                                                         Std Error          Design
                                  race    Variable            Mean         of Mean          Effect
                          ------------------------------------------------------------------------
                                     1    HI_CHOL         0.101492        0.006246        1.082734
                                     2    HI_CHOL         0.121649        0.006604        1.407822
                                     3    HI_CHOL         0.078640        0.010385        2.091156
                                     4    HI_CHOL         0.099679        0.024666        3.098290
                          ------------------------------------------------------------------------
```

## Python

```{python}
import pandas as pd
from samplics import TaylorEstimator
from samplics.utils.types import PopParam

nhanes = pd.read_csv("../data/nhanes.csv")

nhanes_design_kwargs = dict(
    psu=nhanes["SDMVPSU"],
    stratum=nhanes["SDMVSTRA"],
    samp_weight=nhanes["WTMEC2YR"],
    remove_nan=True,
)

# Mean of HI_CHOL
mean_estimator = TaylorEstimator(PopParam.mean)
mean_estimator.estimate(nhanes["HI_CHOL"], **nhanes_design_kwargs)
hi_chol_means = mean_estimator.to_dataframe()

# Sum of HI_CHOL
total_estimator = TaylorEstimator(PopParam.total)
total_estimator.estimate(nhanes["HI_CHOL"], **nhanes_design_kwargs)
hi_chol_totals = total_estimator.to_dataframe()

# Ratio of HI_CHOL / RIAGENDR
ratio_estimator = TaylorEstimator(PopParam.ratio)
ratio_estimator.estimate(
    y=nhanes["HI_CHOL"], x=nhanes["RIAGENDR"], **nhanes_design_kwargs
)
hi_chol_ratio = ratio_estimator.to_dataframe()

# Proportion of different AGECAT values
prop_estimator = TaylorEstimator(PopParam.prop)
prop_estimator.estimate(nhanes["agecat"], **nhanes_design_kwargs)
agecat_prop = prop_estimator.to_dataframe()

# Quantiles of HI_CHOL
# NA

# Domain analysis of mean of HI_CHOL by race, with design effect
mean_estimator = TaylorEstimator(PopParam.mean)
mean_estimator.estimate(
    nhanes["HI_CHOL"],
    **nhanes_design_kwargs,
    domain=nhanes["race"],
    deff=True,  # Design effect param currently has no effect
)
hi_chol_domain_means = mean_estimator.to_dataframe()


ag_dict = agecat_prop.set_index("_level").to_dict()
hc_dict = hi_chol_domain_means.set_index("_domain").to_dict()

print(
    f"""
    Mean of HI_CHOL: {hi_chol_means["_estimate"][0]}
    SE of Mean HI_CHOL: {hi_chol_means["_stderror"][0]}
    CL of Mean HI_CHOL: {(hi_chol_means["_lci"][0], hi_chol_means["_uci"][0])}
    Sum of HI_CHOL: {hi_chol_totals["_estimate"][0]}
    SE of Sum HI_CHOL: {hi_chol_totals["_stderror"][0]}
    CL of Sum HI_CHOL: {(hi_chol_totals["_lci"][0], hi_chol_totals["_uci"][0])}
    Ratio of HI_CHOL / RIAGENDR: {hi_chol_ratio["_estimate"][0]}
    SE of Ratio HI_CHOL / RIAGENDR: {hi_chol_ratio["_stderror"][0]}
    CL of Ratio HI_CHOL / RIAGENDR: {(hi_chol_ratio["_lci"][0], hi_chol_ratio["_uci"][0])}
    Proportion of AGECAT: {ag_dict["_estimate"]}
    SE of Proportion AGECAT: {ag_dict["_stderror"]}
    LCL of Proportion AGECAT: {ag_dict["_lci"]}
    UCL of Proportion AGECAT: {ag_dict["_uci"]}
    Quantiles of HI_CHOL: Not available
    Mean of HI_CHOL by race: {hc_dict["_estimate"]}
    SE of HI_CHL by race: {hc_dict["_stderror"]}
    LCL of HI_CHOL by race: {hc_dict["_lci"]}
    UCL of HI_CHOL by race: {hc_dict["_uci"]}
    Design Effect of HI_CHOL by race: Not available
    """
)
```
:::

# Differences

## Quantiles {#quantiles}

`samplics` in Python does not have a method for calculating quantiles, and in R and SAS the available methods lead to different results. To demonstrate the differences in calculating quantiles, we will use the `apisrs` dataset from the `survey` package in R [@API_2000].

```{r}
#| message: false
library(survey)

data("api")

head(apisrs) |> gt::gt()
```

In SAS, PROC SURVEYMEANS will calculate quantiles of specific probabilities as you request them, using Woodruff's method for intervals and a custom quantile method [@SAS_2018, pp. 9834]. The quantile method does not match any of the available `qrules` in R, and although the default `interval.types` in the R `survey::svyquantile` function also uses Woodruff's method, it is a different implementation.

The method and results from SAS are as follows:

``` default
proc surveymeans data=apisrs total=6194 quantile=(0.025 0.5 0.975);
    var growth;
run;
```

```         
                             The SURVEYMEANS Procedure

                                    Data Summary

                        Number of Observations           200




                                     Quantiles

                                                      Std
 Variable       Percentile       Estimate           Error    95% Confidence Limits
 ---------------------------------------------------------------------------------
 growth           2.5          -16.500000        1.755916    -19.962591 -13.037409
                   50 Median    26.500000        1.924351     22.705263  30.294737
                 97.5           99.000000       16.133827     67.184794 130.815206
 ---------------------------------------------------------------------------------
```

If in R we use the default `qrule="math"` (equivalent to `qrule="hf1"` and matches `type=1` in the `quantile` function for unweighted data) along with the default `interval.type="mean"`, we get the following results:

```{r}
srs_design <- svydesign(data = apisrs,id = ~1,fpc = ~fpc,)

svyquantile(
  ~growth,
  srs_design,
  quantiles = c(0.025, 0.5, 0.975),
  ci=TRUE,
  se=TRUE
)
```

Here we can see that the quantiles, confidence intervals, and standard errors do not match SAS. From testing, none of the available `qrule` methods match SAS for the quantile values, so it is recommended to use the default values unless you have need of some of the other properties of different quantile definitions - see [`vignette("qrule", package="survey")`](https://cran.r-project.org/web/packages/survey/vignettes/qrule.pdf) for more detail. If an exact match to SAS is required, then the `svyquantile` function allows for passing a custom function to the `qrule` argument to define your own method for calculating quantiles. Below is an example that will match SAS:

```{r}
sas_qrule <- function(x, w, p) {
  # Custom qrule to match SAS, based on survey::oldsvyquantile's internal method
  if (any(is.na(x))) 
    return(NA * p)
  w <- rowsum(w, x, reorder = TRUE)
  x <- sort(unique(x))
  cum.w <- cumsum(w)/sum(w)
  cdf <- approxfun(cum.w, x, method = "linear", f = 1, 
    yleft = min(x), yright = max(x), ties = min)
  cdf(p)
}


sas_quants <- svyquantile(
  ~growth,
  srs_design,
  quantiles = c(0.025, 0.5, 0.975),
  qrule=sas_qrule,
  ci=TRUE,
  se=TRUE
)

sas_quants
```

Note that although the quantiles and standard errors match, the confidence intervals still do not match SAS. For this another custom calculation is required, based on the formula used in SAS:

```{r}
sas_quantile_confint <- function(newsvyquantile, level=0.05, df=Inf) {
  q <- coef(newsvyquantile)
  se <- SE(newsvyquantile)
  ci <- cbind(
    q,
    q + se * qt(level/2, df),
    q - se * qt(1 - level/2, df),
    se
  )
  colnames(ci) <- c("quantile", paste0("ci.", c(100 * level / 2, 100 * (1 - level / 2))), "se")

  ci
}

sas_quantile_confint(sas_quants, df=degf(srs_design))
```

## Other considerations

### Degrees of Freedom

Some of the functions in R require the degrees of freedom to be specified when calculating confidence intervals, otherwise it assumes a normal distribution. This can be done easily by using the `survey::degf` function, which calculates the degrees of freedom for a survey design object.

### Single PSU Strata

Although it was not apparent with the examples used here, if there is only one PSU from a stratum then R will by default error, whereas SAS will remove that stratum from the variance calculation. This can be changed in R by setting the `options(survey.lonely.psu="certainty")` to match SAS and have it make no contribution to the variance. In `samplics`, this behaviour can be configured using the `single_psu` argument to the `estimate` method, and can be set to to match SAS using `SinglePSUEst.certainty`. This should be considered carefully however, in R and Python there are additional methods of handling single PSUs that may be more appropriate for your use-case.

### Documentation Differences

One key consideration when choosing a statistical package is the documentation available. In this case, both the `survey` package in R and the survey procedures in SAS have a much more comprehensive set of documentation and examples than `samplics` in Python. This includes both detailed examples, as well as the underlying theory and methods used in the calculations including references to the literature.

# Summary and Recommendations

The `{survey}` package in R and the survey procedures in SAS/STAT both provide similar functionality for calculating survey summary statistics. In most cases in both our tests and others, the results are identical ([@2017_YRBS], [@so2020modelling], [@adamico_2009]). Where differences do occur, primarily in calculating quantiles, the methods in R are more varied and well-documented.

In contrast, the `samplics` package in Python is still early in development, and although it does provide some functionality there are still major limitations in both basic statistics (i.e. quantiles) and in more complex methods that were beyond the scope of this document, and the methods are much less well-documented.

::: {.callout-note collapse="true" title="Session Info"}
```{r}
#| echo: false
si <- sessioninfo::session_info("survey", dependencies = FALSE)
# If reticulate is used, si will include python info. However, this doesn't
# include package info and can't really be adapted (in the same way external can)
# So instead we delete this. (Currently bug with specifying info= multi
# element vector in sessioninfo)
si$python <- NULL
si$external <- structure(list("SAS" = "9.04.01M7P080520"), class = c("external_info", "list"))
si
```

```{python}
#| echo: false
# Although sessioninfo::session_info can report on Python config, it depends on `{reticulate}` which
# may not be used and doesn't include package info. Instead, use the following to replicate
import sys
from importlib.metadata import version

# Add packages to the list here
packages = ["samplics"]

max_len = max(len(pkg) for pkg in packages + ["Python"]) + 3
message = (
  "─ Python configuration ".ljust(79, "─") + "\n"
  + " Python".ljust(max_len) + sys.version + "\n"
  + "\n".join(f" {pkg}".ljust(max_len) + version(pkg) for pkg in packages)
)

print(message)
```
:::
