---
title: "R vs SAS Paired T-Test"
---

# Summary

## Goal

The goal of this comparison is to evaluate whether paired t-tests produce equivalent results in R and SAS for both normal and lognormal data. For normal data, results from `stats::t.test()`, `procs::proc_ttest()`, and SAS `PROC TTEST` are compared directly. For lognormal data, results from a log-transformation / back-transformation approach in R are compared with SAS `PROC TTEST` using the `DIST=LOGNORMAL` option.

## Scope

::::::: columns
:::: {.column width="45%"}
::: {.callout-note appearance="minimal" collapse="false"}
## Methodologies

Paired t-test

Lognormal paired t-test
:::
::::

:::: {.column width="55%"}
::: {.callout-note appearance="minimal" collapse="false"}
## Technical implementations

SAS: `PROC TTEST` with and without the `DIST=LOGNORMAL` option

R: `stats::t.test(paired = TRUE)` and log-transformation / back-transformation approach for lognormal data
:::
::::
:::::::

# Prerequisites

## R packages

```{r}
#| output: false 
#| message: false 
library(tibble) # for example data
library(stats) # t.test()
library(procs) # proc_ttest()
```

## Data

### Normal Paired t-test and Lognormal Paired t-test Sample Data

We use a simulated paired dataset containing systolic blood pressure measurements before and after treatment. The same data are used in the R and SAS examples to compare the log-transformation / back-transformation approach in R with SAS `PROC TTEST` using the `DIST=LOGNORMAL` option.

R:

```{r}
pressure <- tibble::tribble(
  ~SBPbefore, ~SBPafter,
  120, 128,
  124, 131,
  130, 131,
  118, 127,
  140, 132,
  128, 125,
  140, 141,
  135, 137,
  126, 118,
  130, 132,
  126, 129,
  127, 135
)


```

SAS:

``` sas
data pressure;
    input SBPbefore SBPafter @@;
    datalines;
120 128 124 131 130 131 118 127
140 132 128 125 140 141 135 137
126 118 130 132 126 129 127 135
;
run;
```

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
library(procs)
```

# Paired t-test Comparison

The following table shows the types of Paired t-test analysis, the capabilities of each language, and whether or not the results from each language match.

| Analysis | Supported in R | Supported in SAS | Results Match | Notes |
|---------------|---------------|---------------|---------------|---------------|
| Paired t-test, normal data | [Yes](../R/ttest_Paired.html#normal) | [Yes](../SAS/ttest_Paired.html#normal) | [Yes](#normal) | In Base R, use `paired = TRUE` on `t.test()` function |
| Paired t-test, lognormal data | [Yes](../R/ttest_Paired.html#lognormal) | [Yes](../SAS/ttest_Paired.html#lognormal) | [Yes](#lognormal) | Apply a natural log transformation to both paired measurements, perform the paired t-test on the log-transformed data, and exponentiate the mean difference and the confidence limits. |

## Comparison Results

### Normal Data {#normal}

Here is a table of comparison values between `t.test()`, `proc_ttest()`, and SAS `PROC TTEST`:

| Statistic                    | t.test()  | proc_ttest() | PROC TTEST | Match | Notes |
|-----------------|-----------|-----------|-----------|-----------|-----------|
| Degrees of Freedom           | 11        | 11           | 11         | Yes   |       |
| t value                      | -1.089648 | -1.089648    | -1.09      | Yes   |       |
| p value                      | 0.2992    | 0.2992       | 0.2992     | Yes   |       |
| Mean Difference              | -1.8333   | -1.8333      | -1.8333    | Yes   |       |
| Lower 95% CL Mean Difference | -5.5365   | -5.5365      | -5.5365    | Yes   |       |
| Upper 95% CL Mean Difference | 1.8698    | 1.8698       | 1.8698     | Yes   |       |

### Lognormal Data {#lognormal}

For lognormal paired data, a paired t-test can be performed in R by applying a natural log transformation to both paired measurements and conducting the analysis on the log-transformed values. The resulting mean difference and confidence limits can then be exponentiated to obtain a geometric mean ratio and corresponding confidence limits. The table below compares results from `stats::t.test()`, `procs::proc_ttest()`, and SAS `PROC TTEST` with the `DIST=LOGNORMAL` option.

Here is a table of comparison values between `t.test()`, `proc_ttest()`, and SAS `PROC TTEST`:

| Statistic | t.test() | proc_ttest() | PROC TTEST | Match | Notes |
|------------|------------|------------|------------|------------|------------|
| Degrees of Freedom | 11 | 11 | 11 | Yes |  |
| t value | -1.094185 | -1.094185 | -1.09 | Yes |  |
| p value | 0.2973 | 0.2973 | 0.2973 | Yes |  |
| Geometric Mean Ratio | 0.9856 | 0.9856 | 0.9856 | Yes | Back-transformed mean difference |
| Lower 95% CL Mean Ratio | 0.9572 | 0.9572 | 0.9572 | Yes | Back-transformed lower confidence limit |
| Upper 95% CL Mean Ratio | 1.0148 | 1.0148 | 1.0148 | Yes | Back-transformed upper confidence limit |

# Summary and Recommendation

For normal data, the R paired t-test capabilities are comparable to SAS. Comparison between SAS and R show identical results for the datasets tried. The **procs** package `proc_ttest()` function is very similar to SAS in the syntax and output produced. `proc_ttest()` also supports by groups, where `t.test()` does not.

For the lognormal version of the paired t-test, equivalent results can be obtained in R by applying a natural log transformation to the paired measurements, performing the paired t-test on the transformed data, and exponentiating the resulting estimates and confidence limits. Comparison with SAS `PROC TTEST` using the `DIST=LOGNORMAL` option showed that the results matched after back-transformation. Therefore, a separate package is not required for this analysis.

# References

R `t.test()` documentation: <https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/t.test>

R `proc_ttest()` documentation: <https://procs.r-sassy.org/reference/proc_ttest.html>

SAS `PROC TTEST` Paired analysis documentation: <https://documentation.sas.com/doc/en/pgmsascdc/9.4_3.4/statug/statug_ttest_syntax08.htm>
