---
title: "R vs SAS Two-Sample T-Test"
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

## Goal

The goal of this comparison is to evaluate whether two-sample t-tests produce equivalent results in R and SAS for both normal and lognormal data. For normal data, results from `stats::t.test()`, `procs::proc_ttest()`, and SAS `PROC TTEST` are compared directly. For lognormal data, results from a log-transformation / back-transformation approach in R are compared with SAS `PROC TTEST` using the `DIST=LOGNORMAL` option.

## Scope

::::::: columns
:::: {.column width="45%"}
::: {.callout-note appearance="minimal" collapse="false"}
## Methodologies

Two-sample t-test

Lognormal two-sample t-test
:::
::::

:::: {.column width="55%"}
::: {.callout-note appearance="minimal" collapse="false"}
## Technical implementations

SAS: `PROC TTEST` with and without the `DIST=LOGNORMAL` option

R: `stats::t.test()` and log-transformation / back-transformation approach for lognormal data
:::
::::
:::::::

# Prerequisites

## R packages

```{r}
#| output: false 
#| message: false 
library(tibble) # for example data
library(stats) # t.test()
library(procs) # proc_ttest()
```

## Data

### Normal two-sample t-test and Lognormal two-sample t-test Sample Data

We use a simulated two-sample dataset containing weight gain measurements for two treatment groups. The same data are used in the R and SAS examples to compare the log-transformation / back-transformation approach in R with SAS `PROC TTEST` using the `DIST=LOGNORMAL` option.

R:

```{r}
d1 <- tibble::tribble(
  ~trt_grp,	~WtGain,
  "placebo",	94, "placebo",	12, "placebo",	26, "placebo",	89,
  "placebo",	88, "placebo",	96, "placebo",	85, "placebo",	130,
  "placebo",	75, "placebo",	54, "placebo",	112, "placebo",	69,
  "placebo",	104, "placebo",	95, "placebo",	53, "placebo",	21,
  "treatment",	45, "treatment",	62, "treatment",	96, "treatment",	128,
  "treatment",	120, "treatment",	99, "treatment",	28, "treatment",	50,
  "treatment",	109, "treatment",	115, "treatment",	39, "treatment",	96,
  "treatment",	87, "treatment",	100, "treatment",	76, "treatment",	80
)
```

SAS:

``` sas
data d1;
    length trt_grp $ 9;
    input trt_grp $ WtGain @@;
    datalines;
placebo    94 placebo    12 placebo    26 placebo    89 
placebo    88 placebo    96 placebo    85 placebo   130 
placebo    75 placebo    54 placebo   112 placebo    69 
placebo   104 placebo    95 placebo    53 placebo    21 
treatment  45 treatment  62 treatment  96 treatment 128 
treatment 120 treatment  99 treatment  28 treatment  50 
treatment 109 treatment 115 treatment  39 treatment  96 
treatment  87 treatment 100 treatment  76 treatment  80 
;

run;
```

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
library(procs)
```

# Two Sample t-test Comparison

The following table shows the types of Two Sample t-test analysis, the capabilities of each language, and whether or not the results from each language match.

| Analysis | Supported in R | Supported in SAS | Results Match | Notes |
|---------------|---------------|---------------|---------------|---------------|
| Two sample Student's t-test | [Yes](../R/ttest_2Sample.html#baseS) | [Yes](../SAS/ttest_2Sample.html#sas) | [Yes](#student) | In Base R, use `t.test()` function with `paired = FALSE` and `var.equal = TRUE` |
| Two sample Welch's t-test | [Yes](../R/ttest_2Sample.html#baseW) | [Yes](../SAS/ttest_2Sample_W.html#sas) | [Yes](#welch) | In Base R, use `t.test()` function with `paired = FALSE` and `var.equal = FALSE` |
| Two sample Lognormal t-test | [Yes](../R/ttest_2Sample.html#baseWlog) | [Yes](../SAS/ttest_2Sample_lognormal.html#sas) | [Yes](#lognormal) |  |

## Comparison Results

### Student's T-Test {#student}

Here is a table of comparison values between `t.test()`, `proc_ttest()`, and SAS `PROC TTEST`:

| Statistic                    | t.test() | proc_ttest() | PROC TTEST | Match | Notes |
|------------|------------|------------|------------|------------|------------|
| Degrees of Freedom           | 30       | 30           | 30         | Yes   |       |
| t value                      | -0.6969  | -0.70        | -0.70      | Yes   |       |
| p value                      | 0.4912   | 0.4912       | 0.4912     | Yes   |       |
| Mean Difference              | -7.9375  | -7.9375      | -7.9375    | Yes   |       |
| Lower 95% CL Mean Difference | -31.1984 | -31.1984     | -31.1984   | Yes   |       |
| Upper 95% CL Mean Difference | 15.3234  | 15.3234      | 15.3234    | Yes   |       |

### Welch's T-Test {#welch}

In the Welch T-test the variance and effective degrees of freedom are calculated using Satterthwaite method.

Here is a table of comparison values between `t.test()`, `proc_ttest()`, and SAS `PROC TTEST` for this example.

Example with unequal variances:

| Statistic                    | t.test() | proc_ttest() | PROC TTEST() | Match | Notes |
|------------|------------|------------|------------|------------|------------|
| Degrees of Freedom           | 29.694   | 29.694       | 29.694       | Yes   |       |
| t value                      | -0.6969  | -0.70        | -0.70        | Yes   |       |
| p value                      | 0.4913   | 0.4913       | 0.4913       | Yes   |       |
| Mean Difference              | -7.9375  | -7.9375      | -7.9375      | Yes   |       |
| Lower 95% CL Mean Difference | -31.2085 | -31.2085     | -31.2085     | Yes   |       |
| Upper 95% CL Mean Difference | 15.3335  | 15.3335      | 15.3335      | Yes   |       |

### Lognormal Data {#lognormal}

For lognormal data for two independent groups, a two-sample t-test can be performed in R by applying a natural log transformation to the measurements in each group and conducting the analysis on the log-transformed values. The resulting mean difference and confidence limits can then be exponentiated to obtain a geometric mean ratio and corresponding confidence limits. The table below compares results from `stats::t.test()`, `procs::proc_ttest()`, and SAS `PROC TTEST` with the `DIST=LOGNORMAL` option.

Here is a table of comparison values between `t.test()`, `proc_ttest()`, and SAS `PROC TTEST`:

| Statistic | t.test() | proc_ttest() | PROC TTEST | Match | Notes |
|------------|------------|------------|------------|------------|------------|
| Degrees of Freedom | 30 | 30 | 30 | Yes |  |
| t value | -0.8763 | -0.88 | -0.88 | Yes |  |
| p value | 0.3878 | 0.3878 | 0.3878 | Yes |  |
| Geometric Mean Ratio | 0.8381 | 0.8381 | 0.8381 | Yes | Back-transformed mean difference |
| Lower 95% CL Mean Ratio | 0.5553 | 0.5553 | 0.5553 | Yes | Back-transformed lower confidence limit |
| Upper 95% CL Mean Ratio | 1.265 | 1.2649 | 1.2649 | Yes | Back-transformed upper confidence limit |

# Summary and Recommendation

For the Student's T-Test, the R two-sample `t.test()` and **procs** `proc_ttest()` capabilities are comparable to SAS. Comparison between SAS and R show identical results for the datasets tried. The **procs** package `proc_ttest()` function is very similar to SAS in the syntax and output produced. The `proc_ttest()` also supports by groups, where `t.test()` does not.

Likewise, for the Welch's T-Test, the R two-sample `t.test()` and **procs** `proc_ttest()` capabilities are comparable to SAS. Comparison between SAS and R show identical results for the datasets tried.

For the lognormal version of the two-sample t-test, equivalent results can be obtained in R by applying a natural log transformation to the measurements in each group, performing the two-sample t-test on the transformed data, and exponentiating the resulting estimates and confidence limits. Comparison with SAS `PROC TTEST` using the `DIST=LOGNORMAL` option showed that the results matched after back-transformation. Therefore, a separate package is not required for this analysis.

# References

R `t.test()` documentation: <https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/t.test>

R `proc_ttest()` documentation: <https://procs.r-sassy.org/reference/proc_ttest.html>

SAS `PROC TTEST` Two-Sample analysis documentation: <https://documentation.sas.com/doc/en/pgmsascdc/9.4_3.4/statug/statug_ttest_examples01.htm>
