---
title: "R vs SAS ANCOVA"
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

## Goal

The goal of this comparison is to evaluate whether ANCOVA produces equivalent results in R and SAS using the same dataset and model specification. Results from the R `stats::lm() function`, the `sasLM::GLM()` function, and SAS `PROC GLM` are compared for sums of squares and least-squares means.

## Scope

::::::: columns
:::: {.column width="45%"}
::: {.callout-note appearance="minimal" collapse="false"}
## Methodologies

- Analysis of covariance (ANCOVA)

- Type I sums of squares

- Type III sums of squares

- Least-squares means
:::
::::

:::: {.column width="55%"}
::: {.callout-note appearance="minimal" collapse="false"}
## Technical implementations

- SAS: `PROC GLM`
- R: `stats::lm()`
- R: `sasLM::GLM()` with `EMEANS=TRUE`
:::
::::
:::::::

# Prerequisites

## R packages

```{r}
#| output: false 
#| message: false 
library(stats)   # lm()
library(sasLM)   # GLM()
```

## Data

The same dataset is used for the R and SAS analyses.

R:

```{r}
df_sas <- tibble( 
  drug = factor(c( "A", "A", "A", "A", "A", "A", "A", "A", "A", "A", 
                   "D", "D", "D", "D", "D", "D", "D", "D", "D", "D", 
                   "F", "F", "F", "F", "F", "F", "F", "F", "F", "F" 
                   )), 
  pre = c( 
    11, 8, 5, 14, 19, 6, 10, 6, 11, 3, 
    6, 6, 7, 8, 18, 8, 19, 8, 5, 15, 
    16, 13, 11, 9, 21, 16, 12, 12, 7, 12 
    ), 
  post = c( 6, 0, 2, 8, 11, 4, 13, 1, 8, 0, 
            0, 2, 3, 1, 18, 4, 14, 9, 1, 9, 
            13, 10, 18, 5, 23, 12, 5, 16, 1, 20 
            ) 
  )
```

SAS:

``` sas
data df_sas;
    input drug $ pre post;
    datalines;
A 11 6
A 8 0
A 5 2
A 14 8
A 19 11
A 6 4
A 10 13
A 6 1
A 11 8
A 3 0
D 6 0
D 6 2
D 7 3
D 8 1
D 18 18
D 8 4
D 19 14
D 8 9
D 5 1
D 15 9
F 16 13
F 13 10
F 11 18
F 9 5
F 21 23
F 16 12
F 12 5
F 12 16
F 7 1
F 12 20
;
run;
```

The ANCOVA model includes drug as the categorical treatment variable and pre as the continuous covariate.

The general ANCOVA model is:

$$ \text{post} = \text{drug} + \text{pre} $$

where:

- **post** is the response variable

- **drug** is the treatment variable

- **pre** is the continuous covariate.

## ANCOVA Comparison

The following table shows the types of One Sample t-test analysis, the capabilities of each language, and whether or not the results from each language match.

| Analysis | Supported in R | Supported in SAS | Results Match | Notes |
|----|----|----|----|----|
| ANCOVA using general linear model and lsmeans | [Yes](../R/ancova.html) | [Yes](../SAS/ancova.html) | [Yes](#comp) | `GLM()` function from **sasLM** with `EMEANS=TRUE` is the easiest to use and matches SAS |
| Type I Sum of Squares | [Yes](#comp) | [Yes](#comp) | [Yes](#comp) | Sequential sums of squares |
| Type II Sum of Squares | [Yes](#comp) | [Yes](#comp) | [Yes](#comp) | `sasLM::GLM()` provides SAS-style output |
| Type III Sum of Squares | [Yes](#comp) | [Yes](#comp) | [Yes](#comp) | `sasLM::GLM()` with `EMEANS=TRUE` matches SAS `LSMEANS` |

## Comparison Results {#comp}

Here is a table of comparison values between `lm()` from the **stats** package, `GLM()` from the **sasLM** package, and SAS `PROC GLM`:

| Statistic              | lm()     | GLM()     | PROC GLM  | Match | Notes |
|------------------------|----------|-----------|-----------|-------|-------|
| Type I, Sum sq, drug   | 293.6000 | 293.6000  | 293.6000  | Yes   |       |
| Type I, Sum sq, pre    | 577.897  | 577.8974  | 577.8974  | Yes   |       |
| Type III, Sum sq, drug | 68.554   | 68.55371  | 68.55371  | Yes   |       |
| Type III, Sum sq, pre  | 577.897  | 577.89740 | 577.89740 | Yes   |       |
| LSmean drugA           | 6.71     | 6.714963  | 6.714963  | Yes   |       |
| LSmean drugD           | 6.82     | 6.823935  | 6.823935  | Yes   |       |
| LSmean drugF           | 10.16    | 10.161102 | 10.161102 | Yes   |       |

# Summary and Recommendation

The R ANCOVA analysis is comparable to SAS. Comparison between SAS and R show identical results for the datasets tried. The **sasLM** package `GLM()` function is very similar to SAS in the output produced. You can also match SAS using the **stats** package with the `lm()` function and various modification functions. See the R page for further information.

# References

R `lm()` documentation: <https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/lm>

R `GLM()` documentation: <https://cran.r-project.org/web/packages/sasLM/sasLM.pdf>

SAS `PROC GLM` documentation: <https://documentation.sas.com/doc/en/statug/15.2/statug_glm_syntax01.htm>
