---
title: "<SAS/R/SAS vs R> <method name>"
---

*Italic words are note for you.\
Your choice to use part or all to fit your topic.*

## Data used

```{eval = FALSE}
data ExData; 
    input Var1 $ Var2 ...; 
    datalines; 
[Data Lines]; 
run; 
```

```{r}
exData <- tibble::tribble( 
  ~Var1, ~Var2, 
  "val1", 1, 
  "val2", 2) 
```

*a sentence or two about the data. Consider using a package like {skimr} to help illustrate the data.*

## Example Code using <package name>

*Describe more details about the method.\
If needed, add link to other web page/paper/SAS help methodology page, and list them in reference section.\
Print key results as needed.\
Add explanations about code and results as needed.*

## Example Code using <another package name>

*Add if there is more than one package*

## Comparison

*If you are comparing more than one package or between different languages, consider adding in a table to illustrate some of the differences. When comparing between languages it can be nice to have a table with links to the pages with deeper dive of each language for a given method method like this table from the ANCOVA page:*

| Analysis                                      | Supported in R          | Supported in SAS          | Results Match | Notes                                                                                    |
|---------------|---------------|---------------|---------------|---------------|
| ANCOVA using general linear model and lsmeans | [Yes](../R/ancova.html) | [Yes](../SAS/ancova.html) | [Yes](#comp)  | `GLM()` function from **sasLM** with `EMEANS=TRUE` is the easiest to use and matches SAS |

*But, even if you are just comparing the difference in same language it can be helpful to have a table with values of interest across different packages. Using dynamic content you can add the values to a table without manually putting in each value.\
SAS result is very reliable in many cases.*

## Conclusion

*add a conclusion about which language(s) or package(s) are more recommended.\
This section could have a feature like paper abstract, e.g. about the selection of package or SAS PROC.*

## Reference

*Cite all sources and references used in the analysis.*

::: {.callout-note collapse="true" title="Session Info"}
```{r}
#| echo: false

# List all the packages needed 
si <- sessioninfo::session_info(c(
  #Create a vector of all the packages used in this file 
))
si
```
:::
