---
title: "Accelerated Failure Time Model"
format: html
editor: visual
---

The accelerated failure model is a survival analysis technique used to model the relationship between the event of interest (e.g., time to failure) and a set of predictor variables. It assumes that the covariates have a multiplicative effect on the hazard (rate of failure) function. In other words, the hazard rate is accelerated or decelerated by a factor that depends on the values of the covariates.

## Mathematical Expression

Mathematically, the AFT model can be expressed as:

$\log(T) = X\beta + \varepsilon$

Where:

-   T is the survival time

-   log(T) is the logarithm of the survival time

-   X is a matrix of predictor variables

-   β is a vector of coefficients representing the effects of the predictor variables on the logarithm of the survival time

-   ε is the error term assumed to follow a certain distribution, such as normal, exponential, weibull or log-normal

## Example of AFT model using "Log-Normal Distribution"

```{r}
library(survival)
attach(lung)
# Fit an AFT model using lognormal distribution
model_aft <- survreg(Surv(time, status) ~ age + sex + ph.ecog, data = lung, dist = "lognormal")
# Model summary
summary(model_aft)
```

The summary output will provide the estimated coefficients, standard errors, and p-values for each predictor variable.

## Hazard Ratio Calculation

```{r}
# Compute hazard ratios (exponentiated coefficients)
hazard_ratios <- exp(coef(model_aft))
hazard_ratios
```

## Interpretation

-   For age, hazard ratio $< 1$ indicates that for each one-unit increase in age, the survival time is slowed down by a factor of 0.98 (or a 2% decrease).
-   For sex, hazard ratio $> 1$ indicates that males have 68% higher hazard rate or accelerated survival time.
-   A hazard ratio of ph.ecog $< 1$ suggests a 30% lower hazard rate or decelerated survival time associated with ph.ecog.

## Example of AFT model using "Weibull Distribution"

```{r}
attach(ovarian)

# Fit an AFT model using lognormal distribution
model_aft_wb <- survreg(Surv(futime, fustat) ~ age + resid.ds + rx, data = ovarian, dist = "weibull")
# Model summary
summary(model_aft_wb)
```

## Hazard Ratio Calculation

```{r}
# Compute hazard ratios (exponentiated coefficients)
hazard_ratio <- exp(coef(model_aft_wb))
hazard_ratio
```

## Interpretation

-   For age, a hazard ratio of 0.93 indicates that for each one-unit increase in age, the survival time is decelerated by a factor of 0.93(or a 7% decrease)

-   For residual disease status, a hazard ratio of 0.60 suggests that a decrease in residual disease status is associated with a 40% lower hazard rate or decelerated survival time.

-   a hazard ratio of 1.67 suggests a 67% higher hazard rate or accelerated survival time for patients receiving a different type of radiation therapy (rx = 2) compared to the reference group (rx = 1).

## Conclusion

Only the extreme value distribution of error leads to AFT models unlike Cox Ph model where survival times does not follow any specific distribution.
