---
title: "Ancova"
output: html_document
date: "2023-06-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tibble)
library(dplyr)
library(broom)
library(emmeans)
library(rstatix)
library(gt)

df_sas <- 
  tribble(
    ~drug, ~pre, ~post,
      "A",   11,     6,   
      "A",    8,     0,   
      "A",    5,     2,   
      "A",   14,     8,   
      "A",   19,    11,
      "A",    6,     4,   
      "A",   10,    13,   
      "A",    6,     1,   
      "A",   11,     8,   
      "A",    3,     0,
      "D",    6,     0,   
      "D",    6,     2,   
      "D",    7,     3,   
      "D",    8,     1,   
      "D",   18,    18,
      "D",    8,     4,   
      "D",   19,    14,   
      "D",    8,     9,   
      "D",    5,     1,   
      "D",   15,     9,
      "F",   16,    13,   
      "F",   13,    10,   
      "F",   11,    18,   
      "F",    9,     5,   
      "F",   21,    23,
      "F",   16,    12,   
      "F",   12,     5,   
      "F",   12,    16,   
      "F",    7,     1,   
      "F",   12,    20) %>% 
  mutate(drug = factor(drug))
my_formula <- as.formula(post ~ drug + pre)
```

## Introduction

ANOVA is a statistical method used to compare the means of three or more groups to determine if at least one group mean is significantly different from the others. Please see the [anova.qmd](ANOVA%20page) for more information. ANCOVA is an extension to ANOVA.

ANCOVA (Analysis of Covariance) is a statistical method that compares the means of two or more groups while controlling for one or more continuous covariates. By adjusting for these covariates, ANCOVA helps to reduce potential confounding effects, allowing for a clearer assessment of the main treatment effects. It assumes linear relationships between covariates and the dependent variable, along with normality and homogeneity of variances.

We follow the example from link [Analysis of Covariance](https://documentation.sas.com/doc/en/pgmsascdc/9.4_3.4/statug/statug_glm_examples04.htm)

## Data Summary

```{r}
df_sas %>% glimpse()
df_sas %>% summary()
```

## The Model

```{r}
model_ancova <- lm(post ~ drug + pre, data = df_sas)
model_glance <- model_ancova %>% glance()
model_tidy   <- model_ancova %>% tidy()
model_glance %>% gt()
model_tidy   %>% gt()
```

```{r}
model_table <- 
  model_ancova %>% 
  anova() %>% 
  tidy() %>% 
  add_row(term = "Total", df = sum(.$df), sumsq = sum(.$sumsq))
model_table %>% gt()
```

### Type 1

```{r}
df_sas %>%
  anova_test(post ~ drug + pre, type = 1, detailed = TRUE) %>% 
  get_anova_table() %>%
  gt()
```

### Type 2

```{r}
df_sas %>% 
  anova_test(post ~ drug + pre, type = 2, detailed = TRUE) %>% 
  get_anova_table() %>% 
  gt()
```

### Type 3

```{r}
df_sas %>%
  anova_test(post ~ drug + pre, type = 3, detailed = TRUE) %>% 
  get_anova_table() %>% 
  gt()
```

### Least Squares Means

```{r}
model_ancova %>% emmeans::lsmeans("drug") %>% emmeans::pwpm(pvals = TRUE, means = TRUE) 
model_ancova %>% emmeans::lsmeans("drug") %>% plot(comparisons = TRUE)
```

## sasLM Package

The following code performs an ANCOVA analysis using the **sasLM** package. This package was written specifically to replicate SAS statistics. The console output is also organized in a manner that is similar to SAS.

```{r eval=TRUE, echo = TRUE, message=FALSE, warning=FALSE}
library(sasLM)

sasLM::GLM(post ~ drug + pre, df_sas, BETA = TRUE, EMEAN = TRUE)

```

Note that the LSMEANS statistics are produced using the `EMEAN = TRUE` option. The `BETA = TRUE` option is equivalent to the `SOLUTION` option in SAS. See the **sasLM** documentation for additional information.
