---
title: "ANOVA"
---

## Introduction

ANOVA (Analysis of Variance) is a statistical method used to compare the means of three or more groups to determine if at least one group mean is significantly different from the others. It helps to test hypotheses about group differences based on sample data.

The key assumptions include:

-   Independence of observations
-   Normality of the data (the distribution should be approximately normal)
-   Homogeneity of variances (similar variances across groups)

Common types include one-way ANOVA (one independent variable) and two-way ANOVA (two independent variables).

One-way ANOVA tests the effect of a single independent variable on a dependent variable (the grouping factor).

Two-way ANOVA tests the effect of two independent variables on a dependent variable and also examines if there is an interaction between the two independent variables.

### **Getting Started**

To demonstrate the various types of sums of squares, we'll create a data frame called `df_disease` taken from the SAS documentation. The corresponding data can be found [here](https://github.com/PSIAIMS/CAMIS/blob/main/data/sas_disease.csv).

```{r}
#| echo: false
#| include: false
library(tidyverse)
library(emmeans)
library(broom)
library(knitr)

knitr::opts_chunk$set(echo = TRUE)

df_disease <- 
  read_csv(
    file = "../data/sas_disease.csv",
    col_types = cols(drug    = col_factor(),
                     disease = col_factor(),
                     y       = col_double()))
```

### The Model {.unnumbered}

For this example, we're testing for a significant difference in `stem_length` using ANOVA. In R, we're using `lm()` to run the ANOVA, and then using `broom::glance()` and `broom::tidy()` to view the results in a table format.

```{r}
lm_model <- lm(y ~ drug + disease + drug*disease, df_disease)
```

The `glance` function gives us a summary of the model diagnostic values.

```{r}
lm_model %>% 
  glance() %>% 
  pivot_longer(everything())
```

The `tidy` function gives a summary of the model results.

```{r}
lm_model %>% tidy()
```

### The Results {.unnumbered}

You'll see that R print the individual results for each level of the drug and disease interaction. We can get the combined F table in R using the `anova()` function on the model object.

```{r}
lm_model %>% 
  anova() %>% 
  tidy() %>% 
  kable()
```

We can add a `Total` row, by using `add_row` and calculating the sum of the degrees of freedom and sum of squares.

```{r}
lm_model %>%
  anova() %>%
  tidy() %>%
  add_row(term = "Total", df = sum(.$df), sumsq = sum(.$sumsq)) %>% 
  kable()
```

### Sums of Squares Tables {.unnumbered}

Unfortunately, it is not easy to get the various types of sums of squares calculations in using functions from base R. However, the `rstatix` package offers a solution to produce these various sums of squares tables. For each type, you supply the original dataset and model to the. `anova_test` function, then specify the ttype and se `detailed = TRUE`.

#### Type I

```{r}
#| message: false
#| warning: false
df_disease %>% 
  rstatix::anova_test(
    y ~ drug + disease + drug*disease, 
    type = 1, 
    detailed = TRUE) %>% 
  rstatix::get_anova_table() %>% 
  kable()
```

#### Type II {.unnumbered}

```{r}
#| message: false
#| warning: false
df_disease %>% 
  rstatix::anova_test(
    y ~ drug + disease + drug*disease, 
    type = 2, 
    detailed = TRUE) %>% 
  rstatix::get_anova_table() %>% 
  kable()
```

#### Type III {.unnumbered}

```{r}
#| message: false
#| warning: false
df_disease %>% 
  rstatix::anova_test(
    y ~ drug + disease + drug*disease, 
    type = 3, 
    detailed = TRUE) %>% 
  rstatix::get_anova_table() %>% 
  kable()
```

#### Type IV {.unnumbered}

In R there is no equivalent operation to the `Type IV` sums of squares calculation in SAS.
