---
title: "Association Analysis for Count Data Using R"
---

```{r}
#| echo: FALSE
#| include: FALSE
library(tidyverse)
```

The most commonly used association analysis methods for count data/contingency tables compare observed frequencies with those expected under the assumption of independence:

$$
X^2 = \sum_{i=1}^k \frac{(x_i-e_i)^2}{e_i}, 
$$ where $k$ is the number of contingency table cells.

-   Chi-Squared test: test statistic is approximately equal to the theoretical chi-squared distribution. Chi-squared approximation is suitable when the expected values in all cells of the contingency table are above 5.

-   Fisher exact test: test of independence if cell frequencies are small. Although in practice it is employed for small sample sizes, the test is valid for all sample sizes.

Other measures for the correlation of two continuous variables are:

-   Pearson correlation coefficient

-   Spearman correlation coefficient

-   Kendall's rank correlation

# Example: Lung Cancer Data

*Data source:* *Loprinzi CL. Laurie JA. Wieand HS. Krook JE. Novotny PJ. Kugler JW. Bartel J. Law M. Bateman M. Klatt NE. et al. Prospective evaluation of prognostic variables from patient-completed questionnaires. North Central Cancer Treatment Group. Journal of Clinical Oncology. 12(3):601-7, 1994.*

Survival in patients with advanced lung cancer from the North Central Cancer Treatment Group. Performance scores rate how well the patient can perform usual daily activities.

```{r}
library(survival) 
glimpse(lung)
```

# $2\times 2$ Contingency Table

We analyze the association between ECOG performance score (fully active vs any symptomatic patient) as rated by physician and weight loss (see `?lung` for details).

```{r}
lung2 <- survival::lung %>% 
  mutate(
    ecog_grp = factor(ph.ecog > 0, labels = c("fully active", "symptomatic")), 
    wt_grp = factor(wt.loss > 0, labels = c("weight loss", "weight gain"))
  ) 

(tab <- table(x = lung2$ecog_grp, y = lung2$wt_grp))

```

## Chi-Squared test

```{r}
chisq.test(tab)
```

## Fisher Exact Test

For $2 \times 2$ contingency tables, p-values are obtained directly using the hypergeometric distribution.

```{r}
fisher.test(tab)
```

# Large Contingency Tables

In a second example, we analyze the association between ECOG and Karnofky performance scores, both rated by physician (see `?lung` for details).

```{r}
(tab2 <- table(x = lung2$ph.ecog, y = lung2$ph.karno))
```

## Chi-Squared Test

```{r}
chisq.test(tab2)
```

The warning means that the smallest expected frequencies is lower than 5. It is recommended to use the Fisher's exact test in this case.

## Fisher Exact Test

For contingency tables larger than $2 \times 2$, p-values are based on simulations, which might require a lot of time (see `?fisher.test` for details).

```{r}
fisher.test(tab2, simulate.p.value=TRUE)
```
