---
title: "Binomial Test"
---

The statistical test used to determine whether the proportion in a binary outcome experiment is equal to a specific value. It is appropriate when we have a small sample size and want to test the success probability $p$ against a hypothesized value $p_0$.

# Coin flips dataset.

-   We will use coin flips dataset generated from SAS simulation to carry out four binomial tests (Exact test, Wald test, Mid-p adjusted test and Wilson score test). This is to ensure that the proportion value obtained from coin flips dataset is similar for both software rather than simulating in both instances, which leads to different proportion value.

-   We will use the various functions for each test to investigate if the proportion of heads is significantly different from 0.5. Therefore:

$H_0 : p = 0.5$

```{r}
# heads
heads_count <- 520
heads_count

# tails
tails_count <- 480
tails_count

# total
total_flips <- 1000
total_flips
```

## 1. Exact Binomial Test.

```{r}
binom.test(heads_count, total_flips, p = 0.5, conf.level = 0.95)
```

### Results:

The output has a p-value $> 0.05$ (chosen level of significance). Hence, we fail to reject the null hypothesis and conclude that the **coin is fair**.

## 2. Wald(Asymptotic) Test.

```{r}
library(DescTools)
p=0.5
phat<-heads_count/total_flips
phat
z <- (phat - p) / sqrt(p * (1 - p) / total_flips)
2 * (1 - pnorm(abs(z)))
BinomCI(heads_count, total_flips, method = "wald")
```

## Results:

The output has a p-value $> 0.05$ (chosen level of significance). Hence, we fail to reject the null hypothesis and conclude that the **coin is fair**.

## 3. Mid-P adjusted Exact Binomial Test.

```{r}
library(exactci)
binom.exact(heads_count,total_flips, p = 0.5,alternative = "greater", midp = TRUE,tsmethod = "central")
binom.exact(heads_count,total_flips, p = 0.5,alternative = "less", midp = TRUE,tsmethod = "central")
binom.exact(heads_count,total_flips, p = 0.5, midp = TRUE,tsmethod = "central")
```

## Results:

The output has a p-value $> 0.05$ (chosen level of significance). Hence, we fail to reject the null hypothesis and conclude that the **coin is fair**.

## 3. Wilson Score Test.

```{r}
prop.test(heads_count, total_flips, p = 0.5, correct = FALSE)
```

## Results:

The output has a p-value $> 0.05$ (chosen level of significance). Hence, we fail to reject the null hypothesis and conclude that the **coin is fair**.

# Example of Clinical Trial Data.

We load the `lung` dataset from `survival` package. We want to test if the proportion of patients with survival status 1 (dead) is significantly different from a hypothesized proportion (e.g. 50%)

$H_0 : p = 0.19$

We will calculate number of deaths and total number of patients.

```{r}
library(survival)
attach(lung)

# deaths
num_deaths <- sum(lung$status == 1)
num_deaths

# total patients
total_pat <- nrow(lung)
total_pat
```

## 1. Exact Binomial Test.

```{r}
# Exact (Clopper–Pearson)
binom.test(num_deaths, total_pat, p = 0.19, conf.level = 0.95)
```

## Results:

The output has a p-value $< 0.05$ (chosen level of significance). Hence, we reject the null hypothesis and conclude that **the proportion of death is significantly different from 19%**.

## 2. Wald(Asymptotic) Test

```{r}
library(DescTools)
p=0.19
phat<-num_deaths/total_pat
z <- (phat - p) / sqrt(p * (1 - p) / total_pat)
 2 * (1 - pnorm(abs(z)))
BinomCI(num_deaths, total_pat, method = "wald")
```

## Results:

The output has a p-value $< 0.05$ (chosen level of significance). Hence, we reject the null hypothesis and conclude that **the proportion of death is significantly different from 19%**.

## 3. Mid-P adjusted Exact Binomial Test.

```{r}
library(exactci)
binom.exact(num_deaths, total_pat, p = 0.19, midp = TRUE,tsmethod = "central")
binom.exact(num_deaths, total_pat, p = 0.19,alternative = "less", midp = TRUE,tsmethod = "central")
binom.exact(num_deaths, total_pat, p = 0.19,alternative = "greater", midp = TRUE,tsmethod = "central")
```

## Results:

The output for right sided and two sided test has a p-value $< 0.05$ (chosen level of significance). Hence, we reject the null hypothesis and conclude that **the proportion of death is significantly different from 19%**.

## 4. Wilson Score Test

```{r}
prop.test(num_deaths, total_pat, p = 0.19, correct = FALSE)
```

## Results:

The output has a p-value $< 0.05$ (chosen level of significance). Hence, we reject the null hypothesis and conclude that **the proportion of death is significantly different from 19%**.

## Summary:

| Data           | Test                      | P_Value |
|----------------|---------------------------|---------|
| Coin Flips     | Exact Test                | 0.2174  |
|                | Wald Test                 | 0.2059  |
|                | Mid-p adjusted Exact Test | 0.2061  |
|                | Wilson score Test         | 0.2059  |
| Clinical Trial | Exact Test                | 0.0017  |
|                | Wald Test                 | 0.0009  |
|                | Mid-p adjusted Exact Test | 0.0015  |
|                | Wilson Score Test         | 0.0009  |

For the two datasets, the results for Wald and Wilson Score test match. This implies that the sample data are adequate because Wald and Wilson differ mainly when sample size is small or probability of success is close to \[0,1\]. In that case Wilson score test will have better coverage.

