---
title: "Binomial Test"
format: html
editor: visual
---

The statistical test used to determine whether the proportion in a binary outcome experiment is equal to a specific value. It is appropriate when we have a small sample size and want to test the success probability $p$ against a hypothesized value $p_0$.

## Creating a sample dataset

-   We will generate a dataset where we record the outcomes of 1000 coin flips.

-   We will use the `binom.test` function to test if the proportion of heads is significantly different from 0.5.

```{r}
set.seed(19)
coin_flips <- sample(c("H", "T"), 
                     size = 1000, 
                     replace = T,
                     prob = c(0.5, 0.5))
```

Now, we will count the heads and tails and summarize the data.

```{r}
heads_count <- sum(coin_flips == "H")
tails_count <- sum(coin_flips == "T")
total_flips <- length(coin_flips)
```

```{r}
heads_count
tails_count
total_flips
```

## Conducting Binomial Test

```{r}
binom_test_result <- binom.test(heads_count, total_flips, p = 0.5)
binom_test_result
```

### Results:

The output has a p-value `r binom_test_result$p.value` $> 0.05$ (chosen level of significance). Hence, we fail to reject the null hypothesis and conclude that the **coin is fair**.

# Example of Clinical Trial Data

We load the `lung` dataset from `survival` package. We want to test if the proportion of patients with survival status 1 (dead) is significantly different from a hypothesized proportion (e.g. 50%)

We will calculate number of deaths and total number of patients.

```{r}
library(survival)
attach(lung)

num_deaths <- sum(lung$status == 1)
total_pat <- nrow(lung)
```

```{r}
num_deaths
total_pat
```

## Conduct the Binomial Test

We will conduct the Binomial test and hypothesize that the proportin of death should be 19%.

```{r}
binom_test <- binom.test(num_deaths, total_pat, p = 0.19)
binom_test
```

## Results:

The output has a p-value `r binom_test$p.value` $< 0.05$ (chosen level of significance). Hence, we reject the null hypothesis and conclude that **the propotion of death is significantly different from 19%**.
