---
title: "Confidence Intervals for Paired Proportions in R"
---

## Introduction

\[See separate page for general introductory information on confidence intervals for proportions.\]

\[Note: information about cicalc package will be added to this page soon.\]

## Data used

The adcibc data stored [here](../data/adcibc.csv) was used in this example, creating a binary treatment variable `trt` taking the values of `ACT` or `PBO` and a binary response variable `resp` taking the values of `Yes` or `No`. For this example, a response is defined as a score greater than 4.

```{r}
#| echo: false
#| include: false
library(tidyverse)
library(cardx)
library(DescTools)
adcibc2 <- read_csv("../data/adcibc.csv")

adcibc <- adcibc2 |>
  select(AVAL, TRTP) |>
  mutate(
    resp = if_else(AVAL > 4, "Yes", "No"),
    respn = if_else(AVAL > 4, 1, 0),
    trt = if_else(TRTP == "Placebo", "PBO", "ACT"),
    trtn = if_else(TRTP == "Placebo", 0, 1)
  ) |>
  select(trt, trtn, resp, respn)

# cardx package required a vector with 0 and 1s for a single proportion CI
act <- filter(adcibc, trt == "ACT") |>
  select(respn)
act2 <- act$respn
```

The below shows that for the Actual Treatment, there are 36 responders out of 154 subjects = 0.234 (23.4% responders).

```{r}
#| echo: false
adcibc |>
  group_by(trt, resp) |>
  tally()
```

## Packages

**The {ratesci} package** is ... \[TBC\]

**The {ExactCIdiff} package** produces exact CIs for two dependent proportions (matched pairs).

## Methods for Calculating Confidence Intervals for Proportion Difference from matched pairs using {ExactCIdiff} and {ratesci}

For more information about the detailed methods for calculating confidence intervals for a matched pair proportion see [here](https://psiaims.github.io/CAMIS/SAS/ci_for_prop.html#methods-for-calculating-confidence-intervals-for-a-matched-pair-proportion). When you have 2 measurements on the same subject, the 2 sets of measures are not independent and you have matched pair of responses.

To date we have not found an R package which calculates a CI for matched pair proportions using the normal approximation or Wilson methods although they can be done by hand using the equations provided on the SAS page link above.

**The {ExactCIdiff} package** produces exact CIs for two dependent proportions (matched pairs), claiming to be the first package in R to do this method. However, it should only be used when the sample size is not too large as it can be computationally intensive.\
NOTE that the {ExactNumCI} package should not be used for this task. More detail on these two packages can be found [here](RJ-2013-026.pdf).

Using a cross over study as our example, a 2 x 2 table can be formed as follows:

+-----------------------+---------------+---------------+---------------+
|                       | Placebo\      | Placebo\      | Total         |
|                       | Response= Yes | Response = No |               |
+=======================+===============+===============+===============+
| Active Response = Yes | r             | s             | r+s           |
+-----------------------+---------------+---------------+---------------+
| Active Response = No  | t             | u             | t+u           |
+-----------------------+---------------+---------------+---------------+
| Total                 | r+t           | s+u           | N = r+s+t+u   |
+-----------------------+---------------+---------------+---------------+

: The proportions of subjects responding on each treatment are:

Active: $\hat p_1 = (r+s)/n$ and Placebo: $\hat p_2= (r+t)/n$

Difference between the proportions for each treatment are: $D=p1-p2=(s-t)/n$

Suppose :

+-----------------------+---------------+---------------+------------------+
|                       | Placebo\      | Placebo\      | Total            |
|                       | Response= Yes | Response = No |                  |
+=======================+===============+===============+==================+
| Active Response = Yes | r = 20        | s = 15        | r+s = 35         |
+-----------------------+---------------+---------------+------------------+
| Active Response = No  | t = 6         | u = 5         | t+u = 11         |
+-----------------------+---------------+---------------+------------------+
| Total                 | r+t = 26      | s+u = 20      | N = r+s+t+u = 46 |
+-----------------------+---------------+---------------+------------------+

Active: $\hat p_1 = (r+s)/n$ =35/46 =0.761 and Placebo: $\hat p_2= (r+t)/n$ = 26/46 =0.565

Difference = 0.761-0.565 = 0.196, then PairedCI() function can provide an exact confidence interval as shown below

-0.00339 to 0.38065

```{r}
#| eval: false
# ExactCIdiff::PairedCI(s, r+u, t, conf.level = 0.95)

CI <- ExactCIdiff::PairedCI(15, 25, 6, conf.level = 0.95)$ExactCI
CI
```

## Methods for Calculating Confidence Intervals for Relative Risk from matched pairs using {ratesci}

\[TBC\]

## Methods for Calculating Confidence Intervals for Conditional Odds Ratio from matched pairs using {ratesci}

\[TBC\]

## Continuity Adjusted Methods

\[TBC\]

## Consistency with Hypothesis Tests

\[TBC\] - cf. McNemar test

##  

##  

## References

1.  [pharmaverse cardx package](https://insightsengineering.github.io/cardx/main/#:~:text=The%20%7Bcardx%7D%20package%20is%20an%20extension%20of%20the,Data%20Objects%20%28ARDs%29%20using%20the%20R%20programming%20language.)
2.  [PropCIs package](https://cran.r-project.org/web//packages/PropCIs/PropCIs.pdf)
3.  D. Altman, D. Machin, T. Bryant, M. Gardner (eds). Statistics with Confidence: Confidence Intervals and Statistical Guidelines, 2nd edition. John Wiley and Sons 2000.
4.  <https://www.lexjansen.com/wuss/2016/127_Final_Paper_PDF.pdf>
