---
title: "Confidence Intervals for Proportions in R"
---

## Introduction

\[See separate page for general introductory information on confidence intervals for proportions.\]

\[Note: information about cicalc package will be added to this page soon.\]

## Data used

The adcibc data stored [here](../data/adcibc.csv) was used in this example, creating a binary treatment variable `trt` taking the values of `ACT` or `PBO` and a binary response variable `resp` taking the values of `Yes` or `No`. For this example, a response is defined as a score greater than 4.

```{r}
#| echo: false
#| include: false
library(tidyverse)
library(cardx)
library(DescTools)
adcibc2 <- read_csv("../data/adcibc.csv")

adcibc <- adcibc2 |>
  select(AVAL, TRTP) |>
  mutate(
    resp = if_else(AVAL > 4, "Yes", "No"),
    respn = if_else(AVAL > 4, 1, 0),
    trt = if_else(TRTP == "Placebo", "PBO", "ACT"),
    trtn = if_else(TRTP == "Placebo", 0, 1)
  ) |>
  select(trt, trtn, resp, respn)

# cardx package required a vector with 0 and 1s for a single proportion CI
act <- filter(adcibc, trt == "ACT") |>
  select(respn)
act2 <- act$respn
```

The below shows that for the Actual Treatment, there are 36 responders out of 154 subjects = 0.234 (23.4% responders).

```{r}
#| echo: false
adcibc |>
  group_by(trt, resp) |>
  tally()
```

## Packages

**The {cardx} package** is an extension of the {cards} package, providing additional functions to create Analysis Results Data Objects (ARDs)^1^. It was developed as part of {NEST} and pharmaverse. This package requires the binary endpoint to be a logical (TRUE/FALSE) vector or a numeric/integer coded as (0, 1) with 1 (TRUE) being the success you want to calculate the confidence interval for.

See [here](R:%20Functions%20for%20Calculating%20Proportion%20Confidence%20Intervals) for full description of the {cardx} proportions equations.

If calculating the CI for a difference in proportions, the package requires both the response and the treatment variable to be numeric/integer coded as (0, 1) (or logical vector).

Instead of the code presented below, you can use `ard_categorical_ci(data, variables=resp, method ='wilson')` for example. This invokes the code below but returns an analysis results dataset (ARD) format as the output. Methods included are waldcc, wald, clopper-pearson, wilson, wilsoncc, strat_wilson, strat_wilsoncc, agresti-coull and jeffreys for one-sample proportions and methods for 2 independent samples, however currently does not have a method for 2 matched proportions.

**The {PropCIs} package** produces CIs for methods such as Blaker's exact method and Midp which aren't available in {cardx} but are available in SAS. We found results agreed with SAS to the 5th decimal place. The package also calculates CIs for Clopper-Pearson, Wald, Wilson, Agresti-Coull and these align to results obtained in cardx to at least the 7th decimal place. The {PropsCIs} package requires just the number of events (numerator number of successes) & total number of subjects (denominator) as an input dataset. Given Blaker and Midp are rarely used in practice, and {PropsCIs} isn't a package commonly downloaded from CRAN, further details are not provided here.

**The {Hmisc} package** produces CIs using the Clopper-Pearson method. In this example (x=36 and n=154), the results match the cardx package. Documentation reports that the method uses F distribution to compute exact intervals based on the binomial cdf. However, if the percentage of responders is 100% then the upper limit is set to 1. Similarly if the percentage of responders is 0%, then the lower limit is set to 0. Hence, in extreme cases there may be differences between this package and the standard implementation of Clopper-Pearson method.

**The {RBesT} package (Prior to Version 1.8-0)** produces CIs using the Clopper-Pearson method. In this example (x=36 and n=154), the results match the cardx package. However, as described below, there are 2 cases where the results using RBesT package do not match cardx or Hmisc.

1)  x = 0 (0% responders), in which case the lower limit does not match.
2)  x = n (100% responders), in which case the upper limit does not match.

Because of the relationship between the binomial distribution and the beta distribution. This package uses quantiles of the beta distribution to derive exact confidence intervals.

$$ B(\alpha/2;x, n-x+1) < p < B(1-\alpha/2; x+1, n-x)$$

RBesT equations are:\
pLow \<- qbeta(Low, r + (r == 0), n - r + 1)\
pHigh \<- qbeta(High, r + 1, n - r + ((n - r) == 0))

In Version 1.8-0 onwards the equations were updated as follows, which then match the Hmisc intervals:\
pLow \<- qbeta(Low, r, n - r + 1)\
pHigh \<- qbeta(High, r + 1, n - r)

**The {ExactCIdiff} package** produces exact CIs for two dependent proportions (matched pairs).

**The {DescTools} package** has a function BinomDiffCI which produces CIs for two independent proportions (unmatched pairs) including methods for Agresti/Caffo, Wald, Wald with Continuity correction, Newcombe Score, Newcombe score with continuity correction, and more computationally intensive methods such as Miettinen and Nurminen, Mee, Brown Li's Jeffreys, Hauck-Anderson and Haldane. See [here](https://search.r-project.org/CRAN/refmans/DescTools/html/BinomDiffCI.html) for more detail.

**The {presize} package** has a function prec_prop() which also calculates CIs for 2 independent samples using the Wilson, Agresti-Coull, Exact or Wald approaches. The package is not described in further detail here since in most cases **{DescTools}** will be able to compute what is needed. However, it's mentioned due to other functionality it has available such as sample size and precision calculations for AUC, correlations, cronbach's alpha, intraclass correlation, Cohen's kappa, likelihood ratios, means, mean differences, odds ratios, rates, rate ratios, risk differences and risk ratios.

## Methods for Calculating Confidence Intervals for a matched pair proportion using {ExactCIdiff}

For more information about the detailed methods for calculating confidence intervals for a matched pair proportion see [here](https://psiaims.github.io/CAMIS/SAS/ci_for_prop.html#methods-for-calculating-confidence-intervals-for-a-matched-pair-proportion). When you have 2 measurements on the same subject, the 2 sets of measures are not independent and you have matched pair of responses.

To date we have not found an R package which calculates a CI for matched pair proportions using the normal approximation or Wilson methods although they can be done by hand using the equations provided on the SAS page link above.

**The {ExactCIdiff} package** produces exact CIs for two dependent proportions (matched pairs), claiming to be the first package in R to do this method. However, it should only be used when the sample size is not too large as it can be computationally intensive.\
NOTE that the {ExactNumCI} package should not be used for this task. More detail on these two packages can be found [here](RJ-2013-026.pdf).

Using a cross over study as our example, a 2 x 2 table can be formed as follows:

+-----------------------+---------------+---------------+---------------+
|                       | Placebo\      | Placebo\      | Total         |
|                       | Response= Yes | Response = No |               |
+=======================+===============+===============+===============+
| Active Response = Yes | r             | s             | r+s           |
+-----------------------+---------------+---------------+---------------+
| Active Response = No  | t             | u             | t+u           |
+-----------------------+---------------+---------------+---------------+
| Total                 | r+t           | s+u           | N = r+s+t+u   |
+-----------------------+---------------+---------------+---------------+

: The proportions of subjects responding on each treatment are:

Active: $\hat p_1 = (r+s)/n$ and Placebo: $\hat p_2= (r+t)/n$

Difference between the proportions for each treatment are: $D=p1-p2=(s-t)/n$

Suppose :

+-----------------------+---------------+---------------+------------------+
|                       | Placebo\      | Placebo\      | Total            |
|                       | Response= Yes | Response = No |                  |
+=======================+===============+===============+==================+
| Active Response = Yes | r = 20        | s = 15        | r+s = 35         |
+-----------------------+---------------+---------------+------------------+
| Active Response = No  | t = 6         | u = 5         | t+u = 11         |
+-----------------------+---------------+---------------+------------------+
| Total                 | r+t = 26      | s+u = 20      | N = r+s+t+u = 46 |
+-----------------------+---------------+---------------+------------------+

Active: $\hat p_1 = (r+s)/n$ =35/46 =0.761 and Placebo: $\hat p_2= (r+t)/n$ = 26/46 =0.565

Difference = 0.761-0.565 = 0.196, then PairedCI() function can provide an exact confidence interval as shown below

-0.00339 to 0.38065

```{r}
#| eval: false
# ExactCIdiff::PairedCI(s, r+u, t, conf.level = 0.95)

CI <- ExactCIdiff::PairedCI(15, 25, 6, conf.level = 0.95)$ExactCI
CI
```

## References

1.  [pharmaverse cardx package](https://insightsengineering.github.io/cardx/main/#:~:text=The%20%7Bcardx%7D%20package%20is%20an%20extension%20of%20the,Data%20Objects%20%28ARDs%29%20using%20the%20R%20programming%20language.)
2.  [PropCIs package](https://cran.r-project.org/web//packages/PropCIs/PropCIs.pdf)
3.  D. Altman, D. Machin, T. Bryant, M. Gardner (eds). Statistics with Confidence: Confidence Intervals and Statistical Guidelines, 2nd edition. John Wiley and Sons 2000.
4.  <https://www.lexjansen.com/wuss/2016/127_Final_Paper_PDF.pdf>
