---
title: "Confidence Intervals for Independent Proportions in R"
bibliography: references.bib
csl: nature.csl 
---

# Introduction

This page covers confidence intervals for comparisons of two independent proportions in R, including the contrast parameters for risk difference (RD) $\theta_{RD} = p_1 - p_2$, relative risk (RR) $\theta_{RR} = p_1 / p_2$, and odds ratio (OR) $\theta_{OR} = p_1(1-p_2) / (p_2(1-p_1))$.

See the [summary page](../method_summary/ci_for_prop_intro.html) for general introductory information on confidence intervals for proportions, including the principles underlying the most common methods.

Note that because the Asymptotic Score methods rely on an iterative root-finding subroutine to identify the confidence limits, the precision of results (and therefore agreement between packages) depends on the tolerance parameter used.

# Data used

The adcibc data stored [here](../data/adcibc.csv) was used in this example, creating a binary treatment variable `trt` taking the values of `ACT` or `PBO` and a binary response variable `resp` taking the values of `Yes` or `No`. For this example, a response is defined as a score greater than 4.

```{r}
#| echo: false
#| include: false
library(tidyverse)
library(cardx)
library(DescTools)
library(ratesci)
library(contingencytables)
options(digits=6)
adcibc2 <- read_csv("../data/adcibc.csv")

adcibc <- adcibc2 |>
  select(AVAL, TRTP) |>
  mutate(
    resp = if_else(AVAL > 4, "Yes", "No"),
    respn = if_else(AVAL > 4, 1, 0),
    trt = if_else(TRTP == "Placebo", "PBO", "ACT"),
    trtn = if_else(TRTP == "Placebo", 0, 1)
  ) |>
  select(trt, trtn, resp, respn)

# cardx package required a vector with 0 and 1s for a single proportion CI
act <- filter(adcibc, trt == "ACT") |>
  select(respn)
act2 <- act$respn
```

The below shows that for the Active Treatment, there are 36 responders out of 154 subjects, $p_1$ = 0.2338 (23.38% responders), while for the placebo treatment $p_2$ = 12/77 = 0.1558, giving a risk difference of 0.0779, relative risk 1.50, and odds ratio 1.6525.

```{r}
#| echo: false
adcibc |>
  group_by(trt, resp) |>
  tally()
```

# Packages

The table below indicates which methods can be produced using each package, for each contrast. Methods are grouped by those that aim to achieve the nominal confidence interval on average, then the 'exact' and continuity adjusted methods that aim to achieve the nominal confidence level as a minimum. {ExactCIdiff} appears to be the only package offering an 'exact' method for RD, but run times can be prohibitively long, and it is not clear which of the SAS 'exact' methods it matches with, if any.

|   | ratesci | contingencytables | DescTools | cicalc | gsDesign | PropCIs | cardx |
|---------|---------|---------|---------|---------|---------|---------|---------|
| For proximate coverage: |  |  |  |  |  |  |  |
| **Wald/log** | RD,RR,OR | RD,RR,OR | RD,RR,OR | RD | \- | \- | RD |
| **Agresti-Caffo** | RD | RD | RD | \- | \- | \- | \- |
| **MOVER-Wilson (Newcombe)** | RD,RR,OR | RD,RR,OR | RD | RD | \- | \- | \- |
| **MOVER-Jeffreys** | RD,RR,OR | \- | \- | \- | \- | \- | \- |
| Asymptotic Score methods: |  |  |  |  |  |  |  |
| **Miettinen-Nurminen** | RD,RR,OR | RD,RR,OR | RD | RD | RD,RR,OR | RD | \- |
| **Mee/ Koopman** | RD,RR,OR | RD,RR,OR | RD,RR | RD | RD,RR,OR | RR | \- |
| **SCAS** | RD,RR,OR | \- | \- | \- | \- | \- | \- |
| For conservative coverage: |  |  |  |  |  |  |  |
| **Wald-cc** | RD | RD | RD | RD | \- | \- | RD |
| **MOVER-Wilson-cc** | RD,RR,OR | \- | RD | RD | \- | \- | \- |
| **MOVER-Jeffreys-cc** | RD,RR,OR | \- | \- | \- | \- | \- | \- |
| **MN-cc** | RD,RR,OR | \- | \- | \- | \- | \- | \- |
| **SCAS-cc** | RD,RR,OR | \- | \- | \- | \- | \- | \- |

**The {ratesci} package** has the most extensive coverage of the best-performing methods (Asymptotic Score and MOVER) for all contrasts, including several features not available elsewhere, including the skewness correction for improved symmetrical one-sided coverage (SCAS), corresponding hypothesis tests, an option to apply the MOVER approach with Jeffreys intervals (MOVER-J), and optional 'sliding scale' continuity adjustments across all methods. It also produces a selection of other methods for reference.

**The {DescTools} package** has a function `BinomDiffCI()` which produces CIs for RD using numerous different methods including those indicated above plus Brown-Li(Jeffreys), Hauck-Anderson, and Haldane. See [here](https://search.r-project.org/CRAN/refmans/DescTools/html/BinomDiffCI.html) for more detail. The `BinomRatioCI()` function offers several methods for RR CIs, including the approximate (log)normal and Asymptotic Score (Koopman) methods. The methods available for OR are more limited: `OddsRatio()` only has Wald, MLE and mid-P CI methods.

**The {PropCIs} package** has functions `diffscoreci()` for the Miettinen-Nurminen (MN) CI for RD, `riskscoreci()` for the Koopman interval for RR, and `orscoreci()` for the MN CI for OR. It also has functions providing Bayesian tail intervals (with user-specified prior).

**The {contingencytables} package** also provides a good selection of different methods.

**The {cicalc} package** in general replicates the methods available in SAS, but only for the RD contrast.

**The {gsDesign} package** gives Asymptotic Score intervals (with or without 'N-1' correction) for all contrasts with the `ciBinomial()` function. The package also has functions for the corresponding hypothesis tests, and sample size calculations.

**The {cardx} package** has very limited options for comparing proportions - the `cardx::ard_stats_prop_test` function only provides for estimation of RD with the Wald approximate normal method (which is not recommended).

```{=html}
<!-- #
Note to previous contributor: description of the {presize} package is removed. The function for RD is called prec_riskdiff, but it doesn't seem very well set up for producing the intervals for a given dataset (user has to specify p1, p2, n1 and allocation ratio). Suggest it doesn't belong in this evaluation.

**The {presize} package** has a function prec_prop() which also calculates CIs for 2 independent samples using the Wilson, Agresti-Coull, Exact or Wald approaches. The package is not described in further detail here since in most cases **{DescTools}** will be able to compute what is needed. However, it's mentioned due to other functionality it has available such as sample size and precision calculations for AUC, correlations, cronbach's alpha, intraclass correlation, Cohen's kappa, likelihood ratios, means, mean differences, odds ratios, rates, rate ratios, risk differences and risk ratios.
-->
```

# Methods for Calculating Confidence Intervals for Proportion Difference from 2 independent samples

This [paper](https://www.lexjansen.com/wuss/2016/127_Final_Paper_PDF.pdf) describes many methods for the calculation of confidence intervals for 2 independent proportions. The 2-sided and 1-sided performance of many of the same methods have been compared graphically[@laud2014]. For more technical information regarding the methods below see the corresponding [SAS page](https://psiaims.github.io/CAMIS/SAS/ci_for_2indep_prop.html).

````{=html}
<!-- #
Note to previous contributor: The section on {cardx} is removed, because the package only offers the inferior Wald method. Note the documentation for prop.test references Newcombe's 1998 paper, which evaluated several methods (including Wald and his proposed hybrid score method) - it does not say which of those methods is implemented, but it clearly is the Wald method.
Also note that the example code for the cardx function calls does not produce readable output for me.

## Normal Approximation Method (Also known as the Wald or asymptotic CI Method) using {cardx}

#### Example code

`cardx::ard_stats_prop_test function` uses `stats::prop.test` which also allows a continuity correction to be applied.

Although this website [here](https://rdrr.io/r/stats/prop.test.html) and this one [here](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/prop.test) both reference Newcombe for the CI that this function uses, replication of the results by hand and compared to SAS show that the results below match the Normal Approximation (Wald method, i.e. 'method 1' in Newcombe's paper).

Both the Treatment variable (ACT,PBO) and the Response variable (Yes,No) have to be numeric (0,1) or logical (TRUE,FALSE) variables.

The prop.test default with 2 groups, is the null hypothesis that the proportions in each group are the same and a 2-sided CI.

```{r}
indat1 <- adcibc2 |>
  select(AVAL, TRTP) |>
  mutate(
    resp = if_else(AVAL > 4, "Yes", "No"),
    respn = if_else(AVAL > 4, 1, 0),
    trt = if_else(TRTP == "Placebo", "PBO", "ACT"),
    trtn = if_else(TRTP == "Placebo", 1, 0)
  ) |>
  select(trt, trtn, resp, respn)

# cardx package required a vector with 0 and 1s for a single proportion CI
# To get the comparison the correct way around Placebo must be 1, and Active 0

indat <- select(indat1, trtn, respn)

cardx::ard_stats_prop_test(
  data = indat,
  by = trtn,
  variables = respn,
  conf.level = 0.95,
  correct = FALSE
)
cardx::ard_stats_prop_test(
  data = indat,
  by = trtn,
  variables = respn,
  conf.level = 0.95,
  correct = TRUE
)
```
-->
````

### Example code for {DescTools}

```{r}
indat1 <- adcibc2 |>
  select(AVAL, TRTP) |>
  mutate(
    resp = if_else(AVAL > 4, "Yes", "No"),
    respn = if_else(AVAL > 4, 1, 0),
    trt = if_else(TRTP == "Placebo", "PBO", "ACT"),
    trtn = if_else(TRTP == "Placebo", 1, 0)
  ) |>
  select(trt, trtn, resp, respn)

# cardx package required a vector with 0 and 1s for a single proportion CI
# To get the comparison the correct way around Placebo must be 1, and Active 0

indat <- select(indat1, trtn, respn)

# BinomDiffCI requires
# x1 = successes in active,  n1 = total subjects in active,
# x2 = successes in placebo, n2 = total subjects in placebo

x <- indat |>
  filter(respn == 1) |>
  count(trtn, respn) |>
  pull(n)
x

n <- indat |>
  count(trtn) |>
  pull(n)
n

DescTools::BinomDiffCI(
  x1 = x[1],
  n1 = n[1],
  x2 = x[2],
  n2 = n[2],
  conf.level = 0.95,
  sides = c("two.sided"),
  method = c(
    "wald",
    "waldcc",
    "score",
    "scorecc",
    "ac",
    "mn",
    "mee",
    "blj",
    "ha",
    "hal",
    "jp"
  )
) |>
  as_tibble(rownames = "Method")
```

### Example code for {ratesci}

```{r}
# Selected methods for proximate coverage
ratesci::rdci(
  x1 = x[1],
  n1 = n[1],
  x2 = x[2],
  n2 = n[2],
  level = 0.95,
  precis = 6
)$estimates 
```

### Example code for {contingencytables}

```{r}

# Tricky to get this the right way round
tab2x2 <- table(indat1$trt, indat1$resp)[, c(2,1)]

contingencytables::the_2x2_table_CIs_difference(n = tab2x2)
```

# Methods for Calculating Confidence Intervals for Relative Risk from 2 independent samples

The 1-sided performance of selected methods have been compared graphically[@laud2017], with the observation that optimum 2-sided coverage follows directly from optimum 1-sided coverage (while the reverse is not true). It has been noted previously that the ratio contrasts suffer a greater imbalance in 1-sided coverage than RD[@gart1990]. Therefore, skewness correction is particularly valuable here.

### Example code for {DescTools}

```{r}
DescTools::BinomRatioCI(
  x1 = x[1],
  n1 = n[1],
  x2 = x[2],
  n2 = n[2],
  conf.level = 0.95,
  sides = c("two.sided"),
  method = c("katz.log", "adj.log", "bailey", "koopman", "noether", 
                        "sinh-1", "boot")
) |>
  as_tibble(rownames = "Method")
```

### Example code for {ratesci}

```{r}
# Selected methods for proximate coverage
ratesci::rrci(
  x1 = x[1],
  n1 = n[1],
  x2 = x[2],
  n2 = n[2],
  level = 0.95,
  precis = 6
)$estimates 
```

### Example code for {gsDesign}

```{r}
# Miettinen-Nurminen
gsDesign::ciBinomial(
  x1 = x[1],
  n1 = n[1],
  x2 = x[2],
  n2 = n[2],
  scale = 'rr', 
  adj = 1
)
```

### Example code for {contingencytables}

```{r}

contingencytables::the_2x2_table_CIs_ratio(n = tab2x2)
```

# Methods for Calculating Confidence Intervals for Odds Ratio from 2 independent samples

### Example code for {ratesci}

```{r}
# Selected methods for proximate coverage
ratesci::orci(
  x1 = x[1],
  n1 = n[1],
  x2 = x[2],
  n2 = n[2],
  level = 0.95,
  precis = 6
)$estimates 
```

### Example code for {gsDesign}

```{r}
# Miettinen-Nurminen
gsDesign::ciBinomial(
  x1 = x[1],
  n1 = n[1],
  x2 = x[2],
  n2 = n[2],
  scale = 'or', 
  adj = 1
)
```

### Example code for {contingencytables}

Note there is an error in the {contingencytables} package v3.1.0 for the MOVER-R Wilson method.

```{r}
contingencytables::the_2x2_table_CIs_OR(n = tab2x2)
```

# Continuity Adjusted Methods

There are relatively few methods widely available for aligning the minimum coverage with the nominal confidence level. The most versatile option is to use functions from the {ratesci} package, which provides optional continuity adjustments, on a sliding scale from 0 to $0.5/N$, for any of the Asymptotic Score or MOVER methods for any contrast.

### Example code for {ratesci}

```{r}
# Selected methods for conservative coverage
# Using the conventional 0.5
ratesci::rdci(
  x1 = x[1],
  n1 = n[1],
  x2 = x[2],
  n2 = n[2],
  level = 0.95,
  cc = TRUE,
  precis = 6
)$estimates 
```

```{r}
# Using an intermediate adjustment of magnitude 0.25
ratesci::rdci(
  x1 = x[1],
  n1 = n[1],
  x2 = x[2],
  n2 = n[2],
  level = 0.95,
  cc = 0.25,
  precis = 6
)$estimates 

```

# Consistency with hypothesis tests

## Test for association

The Asymptotic Score methods for all contrasts are inherently consistent with $\chi^2$ tests. What may be less widely known is that there is more than one version of the $\chi^2$ test. The Mee and Koopman methods (without the 'N-1' variance correction) are consistent with the **Karl** Pearson $\chi^2$ (as produced by `stats::chisq.test()`) , while the Miettinen-Nurminen method agrees with the **Egon** Pearson 'N-1' test. (The stratified MN interval agrees with the standard CMH test, which also incorporates the 'N-1' adjustment.) Note that the SCAS (with or without 'N-1' adjustment) also agrees with the same tests for association, because the skewness correction term is zero when $\theta_{RD}=0$ or when $\theta_{RR}$ or $\theta_{OR}=1$. The 'N-1' adjusted $\chi^2$ test is available in the {ratesci} and {gsDesign} packages.

## Non-inferiority test

One important use for CIs for independent proportions is in the analysis of clinical trials aiming to demonstrate non-inferiority for an outcome such as cure rate or relapse rate. The Asymptotic Score methods are naturally suited for this purpose, as they are derived by inverting a score test statistic. Probably the most well-known named test for such analysis is the Farrington-Manning (FM) test, but it is important to note that the FM formula omits the 'N-1' correction factor, so is consistent with the Mee CI, not the Miettinen-Nurminen CI. Non-inferiority tests including the 'N-1' adjustment can be obtained from the {ratesci} and {gsDesign} functions.

# References
