---
title: "Confidence Intervals for a Proportion in R"
bibliography: references.bib
csl: nature.csl 
---

# Introduction

See the [summary page](../method_summary/ci_for_prop_intro.html) for general introductory information on confidence intervals for proportions, including the principles underlying the most common methods, and a summary of R packages available.

\[Note: information about cicalc package will be added to this page soon.\]

# Data used

The adcibc data stored [here](../data/adcibc.csv) was used in this example, creating a binary treatment variable `trt` taking the values of `ACT` or `PBO` and a binary response variable `resp` taking the values of `Yes` or `No`. For this example, a response is defined as a score greater than 4.

```{r}
#| echo: false
#| include: false
library(tidyverse)
library(cardx)
library(DescTools)
library(ratesci)
adcibc2 <- read_csv("../data/adcibc.csv")

adcibc <- adcibc2 |>
  select(AVAL, TRTP) |>
  mutate(
    resp = if_else(AVAL > 4, "Yes", "No"),
    respn = if_else(AVAL > 4, 1, 0),
    trt = if_else(TRTP == "Placebo", "PBO", "ACT"),
    trtn = if_else(TRTP == "Placebo", 0, 1)
  ) |>
  select(trt, trtn, resp, respn)

# cardx package requires a vector with 0 and 1s for a single proportion CI
act <- filter(adcibc, trt == "ACT") |>
  select(respn)
act2 <- act$respn
```

The below shows that for the Active Treatment, there are 36 responders out of 154 subjects = 0.234 (23.4% responders).

```{r}
#| echo: false
adcibc |>
  group_by(trt, resp) |>
  tally()

```

# Packages

The table below indicates which methods can be produced using each package. Methods are grouped by those that aim to achieve the nominal confidence interval on average, then the 'exact' and continuity adjusted methods that aim to achieve the nominal confidence level as a minimum.

|                            | cardx | PropCIs | ratesci | DescTools | cicalc | Hmisc | RBesT |
|-------------|---------|---------|---------|---------|---------|---------|---------|
| For proximate coverage:    |       |         |         |           |        |       |       |
| **Wald**                   | Y     | Y       | Y       | Y         | Y      | Y     | \-    |
| **Agresti-Coull**          | Y     | Y       | Y       | Y         | Y      | \-    | \-    |
| **Jeffreys**               | Y     | \-      | Y       | Y         | Y      | \-    | \-    |
| **Mid-P**                  | \-    | Y       | Y       | Y         | Y      | \-    | \-    |
| **Wilson**                 | Y     | Y       | Y       | Y         | Y      | Y     | \-    |
| **SCAS**                   | \-    | \-      | Y       | \-        | \-     | \-    | \-    |
| For conservative coverage: |       |         |         |           |        |       |       |
| **Wald-cc**                | Y     | \-      | Y       | Y         | Y      | \-    | \-    |
| **Jeffreys-cc**            | \-    | \-      | Y       | \-        | \-     | \-    | \-    |
| **Clopper-Pearson**        | Y     | Y       | Y       | Y         | Y      | Y     | Y     |
| **Wilson-cc**              | Y     | \-      | Y       | \-        | Y      | \-    | \-    |
| **SCAS-cc**                | \-    | \-      | Y       | \-        | \-     | \-    | \-    |
| **Blaker**                 | \-    | Y       | Y       | Y         | \-     | \-    | \-    |

Most of these packages require just the number of events (numerator number of successes) & total number of subjects (denominator) as an input dataset. The exception is {cardx}, which takes input data as one row per individual.

**The {cardx} package** includes a separate function for each method. Alternatively, you can use `cardx::ard_categorical_ci(data = act, variables = respn, method ='wilson')` for example. This invokes the code for the selected method but returns an analysis results dataset (ARD) format as the output.

**The {PropCIs} package** contains separate functions to produce CIs for various methods, including Blaker's 'exact' method and Mid-P which aren't available in {cardx} but are produced by SAS PROC FREQ. We found results agreed with SAS to the 5th decimal place. For methods given by both {PropCIs} and {cardx}, the results generally aligned to at least the 7th decimal place, but some methods (Wilson & mid-P) are only output with 4 dps.

**The {ratesci} package** (`rateci()` function) produces a variety of CIs for a proportion - note, the current development version at <https://github.com/petelaud/ratesci> has added more methods compared to the CRAN release (v1.0.0). A CRAN update will be released in due course.

**The {DescTools} package** has a function `BinomCI()` that produces a wide selection of methods. It has

**The {cicalc} package** ... \[TBC\]

**In the {Hmisc} package,** `binconf()` produces CIs using either the Clopper-Pearson, Wilson or Wald method. In this example (x=36 and n=154), the results match the cardx package.

```{=html}
<!--#
Note for previous contributor: the following text is removed because there is no contradiction between 'setting the lower limit to 0' when x=0, and 'the standard implementation of the Clopper-Pearson method'. Please explain further (with example) if there are specific cases that lead to a difference.

Documentation reports that the method uses F distribution to compute exact intervals based on the binomial cdf. However, if the percentage of responders is 100% then the upper limit is set to 1. Similarly if the percentage of responders is 0%, then the lower limit is set to 0. Hence, in extreme cases there may be differences between this package and the standard implementation of Clopper-Pearson method. 
-->
```

**The {RBesT} package** produces CIs using the Clopper-Pearson method with the `BinaryExactCI()` function. (Prior to version 1.8-0, the function produced erroneous results for boundary cases where x = 0 or x = N).

Another package producing many of the above intervals is **{contingencytables}**, but the format of its output objects is not very easy to work with.

# Methods for Calculating Confidence Intervals for a single proportion using R

For more technical derivation and characteristics of each of the methods listed below, see the corresponding [SAS page](https://psiaims.github.io/CAMIS/SAS/ci_for_prop.html).

Most of the methods are demonstrated below using the {cardx} package.

## Clopper-Pearson (Exact or binomial CI) Method

The Clopper-Pearson 'Exact' CI is one of the most popular methods - it is often a preferred 'safe option' to avoid the known deficiencies of the Wald method and guarantee coverage of at least the nominal confidence level, but it can be excessively conservative (too wide an interval compared to the interval containing the true population proportion 95% of the time).

The {cardx} package calculates the Clopper-Pearson interval by calling the `stats::binom.test()` function.

```{r}
cardx::proportion_ci_clopper_pearson(act2, conf.level = 0.95) |>
  as_tibble() |>
  select(-(statistic:parameter))
```

## Normal Approximation Method (Also known as the Wald Method)

The following code calculates a confidence interval for a binomial proportion using the normal approximation equation manually. This is replicated exactly using the `cardx::proportion_ci_wald function` which also allows the continuity correction to be applied.

```{r}
# sample proportion by trt
summary <- adcibc |>
  filter(trt == "ACT") |>
  group_by(resp) |>
  tally() |>
  ungroup() |>
  mutate(
    total = sum(n),
    p = n / total
  )

# Calculate standard error and 95% wald confidence intervals for population proportion
waldci <- summary |>
  filter(resp == "Yes") |>
  mutate(
    se = sqrt(p * (1 - p) / total),
    lower_ci = (p - qnorm(1 - 0.05 / 2) * se),
    upper_ci = (p + qnorm(1 - 0.05 / 2) * se)
  )
waldci

# cardx package Wald method without continuity correction
cardx::proportion_ci_wald(act2, conf.level = 0.95, correct = FALSE) |>
  as_tibble()
```

## Wilson Method (Also known as the (Asymptotic) Score method)

The cardx package calculates the Wilson (score) method by calling the `stats::prop.test()` function. This method is often used as a compromise between the Clopper-Pearson and the Wald given it was found to be accurate for most parameter values (even those close to 0 and 1), and it does not suffer from being over-conservative. However, it over-corrects the asymmetric one-sided coverage of Wald - the location of the Wilson interval is shifted too far towards 0.5[@tonycai2005]. For more technical information see the corresponding [SAS page](https://psiaims.github.io/CAMIS/SAS/ci_for_prop.html).

```{r}
# cardx package Wilson method without continuity correction
cardx::proportion_ci_wilson(act2, conf.level = 0.95, correct = FALSE) |>
  as_tibble() |>
  select(-(statistic:parameter))
```

A recent improvement to the Wilson Score method introduces a skewness correction in order to address the asymmetric one-sided coverage probabilities[@laud2017]. The resulting Skewness Corrected Asymptotic Score (SCAS) method can be obtained from the {ratesci} package. Note in this example the interval is shifted downwards by about 0.0015.

```{r}
ratesci::rateci(x = 36, n = 154)$scas |>
  as_tibble()
```

## Agresti-Coull Method

The {cardx} package calculates the Agresti-Coull method using the equation from the published method by Alan Agresti & Brent Coull based on adding 2 successes and 2 failures before computing the wald CI. The CI is truncated, when it overshoots the boundary (\<0 or \>1).

```{r}
# cardx package agresti_coull method
cardx::proportion_ci_agresti_coull(act2, conf.level = 0.95) |>
  as_tibble()
```

## Jeffreys Method

The Jeffreys 'equal-tailed' method is a particular type of Bayesian method, which optimises central location instead of Highest Probability Density (HPD)[@brown2001]. For proportions, the beta distribution is generally used for the prior, which consists of two parameters alpha and beta. Setting alpha=beta=0.5 is called the Jeffreys prior. NOTE: if you want to use any other priors you can use the `ratesci:jeffreysci()` function which allows a different beta prior to be specified using the `ai` and `bi` arguments. Alternatively (and if you prefer to opt for an HPD approach not recommended by Brown et al.), you can use `binom::binom.bayes`.

```{r}
# cardx package jeffreys method
cardx::proportion_ci_jeffreys(act2, conf.level = 0.95) |>
  as_tibble()
```

## Binomial based mid-P Method

The table below (from {ratesci}) includes two versions of the mid-P interval, one derived iteratively from binomial tail proportions and another from quantiles of a beta distribution. These have been claimed to be equivalent[@brown2001], but it seems they are not quite identical - the beta version is very close to the Jeffreys interval. The first version closely matches with the result (0.17200131, 0.30545883) from SAS PROC FREQ:

```{r}
ratesci::rateci(x = 36, n = 154, precis = 10)$estimates[,,1] |>
  as_tibble(rownames = "Method")

```

The mid-P result from {DescTools} doesn't quite match with either of the above results, beyond the 4th decimal place, even with increased precision using the `tol` argument (which according to the documentation only affects the Blaker method). The documentation indicates that the mid-P interval is derived from the F-distribution, which suggests there are actually 3 slightly different versions of 'mid-P'.

```{r}
DescTools::BinomCI(x = 36, n = 154, 
                   tol = 1e-10,
                   method = c("jeffreys", "wilson", "midp", "clopper-pearson")) |> 
  as_tibble(rownames = "Method")

```

The PropCIs version only displays 4dps, because the calculations are performed on sequential values of p at intervals of 0.0005. Therefore the 4th decimal place shown is only ever 1 or 6, and the PropCIs output for this method is only reliable up to the 3rd decimal place. (A similar precision issue is also noted for the `exactci::binom.exact(..., midp = TRUE)` function, which is probably due to reliance on `uniroot()` with its default value for the `tol` argument.)

```{r}
PropCIs::midPci(x = 36, n = 154, conf.level = 0.95) 
```

## Blaker Method

The Blaker 'exact' CI matching SAS output (to at least 7 dps) may be obtained as follows from {ratesci}, or from the dedicated {BlakerCI} package. Output from `DescTools::BinomCI` and `PropCIs::blakerci()` is less precise, matching to 4 decimal places.

```{r}
options(digits=8)
# ratesci package Blaker method 
ratesci::rateci(36, 154, cc = TRUE)$blaker |>
  as_tibble()
```

# Continuity Adjusted Methods

Continuity adjustments (at a magnitude of $\gamma / N$ where conventionally $\gamma = 0.5$ but smaller values may be used) can be applied to the Wald, Wilson and SCAS formulae.

Here is an example of the adjusted Wilson interval from {cardx}:

```{r}
# cardx package Wilson method with continuity correction

cardx::proportion_ci_wilson(act2, conf.level = 0.95, correct = TRUE) |> 
  as_tibble()
```

In principle, the concept can also be applied to the Jeffreys CI, although with $\gamma=0.5$ this turns out to be identical to the Clopper-Pearson interval.

```{r}
ratesci::rateci(x = 36, n = 154, cc = 0.5)$estimates |>
  drop() |>
  as_tibble(rownames="Method")

```

# Consistency with hypothesis tests

The consistency of CIs for a proportion with a hypothesis test depends on what test is specified. Some confidence interval methods are derived by inverting a test, and therefore guarantee consistency. This includes Clopper Pearson (exact binomial test from `prop.test()`), Wilson Score (asymptotic binomial test from `binom.test()` using the standard error estimated under the null hypothesis), and mid-P (exact test with mid-P adjustment, e.g. from `exactci::binom.exact(..., midp = TRUE)`), although given the observations above, it would be important to select the right version of the mid-P interval. For the SCAS method, the CI is consistent with a skewness-corrected version of the asymptotic test (output by the `ratesci::scoreci()` function).

The output from some of the {cardx} functions (e.g. `proportion_ci_wilson()`) and `proportion_ci_clopperpearson()` includes a p-value, which is not mentioned in the package documentation. It is a test against a null hypothesis of p=0.5, which is the default value used by the underlying `binom.test` and `prop.test` functions. If a test against a specified value of p was required, then those underlying functions should be used to obtain the p-value, as there is no facility to select a different null value in the {cardx} functions.

# Stratified Score methods

\[To be moved to separate page for stratified analysis?\]

{cardx} also contains a function `proportion_ci_strat_wilson()` which calculates stratified Wilson CIs for unequal proportions[@yan2010], with or without continuity adjustment, using a combination of the CIs from each stratum. The default weights (not clearly documented) are presumably those described in the publication, which aim to minimise the weighted sum of the squared interval lengths. The function has the facility to specify other weights (such as proportional to stratum size, for example, or adjusted for known population weights). Examples shown use the `adcibc2` dataset, stratified by age group.

```{r}
actstrat <- adcibc2 |>
  select(TRTP, AVAL, AGEGR1) |>
  filter(TRTP != "Placebo") |>
  mutate(
    respn = if_else(AVAL > 4, 1, 0),
    agegrp = factor(AGEGR1)
  ) |>
  select(respn, agegrp)
actage <- actstrat$agegrp

table(actage, act2) |>
  as_tibble()

table(actage) |>
  as_tibble()

cardx::proportion_ci_strat_wilson(x = act2, strata = actage) |>
  as_tibble()

cardx::proportion_ci_strat_wilson(x = act2, strata = actage, weights = c(1/3, 1/3, 1/3)) |>
  as_tibble()
```

The {ratesci} package can also produce stratified Wilson or SCAS intervals (with or without continuity adjustment), though the underlying methodology is different - this version uses a stratified score function[@laud2017] (analogous to the Miettinen-Nurminen formula for stratified risk difference) instead of constructing the interval from a CI calculated separately for each stratum. For stratified analysis, the `ratesci::scoreci()` function takes inputs in the form of vectors of the numerators and denominators per stratum. By default, weights are proportional to stratum size, but custom weights are also catered for using the `wt` argument. Arbitrary fixed weights are shown here to allow comparison with the {cardx} version.

```{r}
ratesci::scoreci(x1 = c(4, 21, 11), 
                 n1 = c(17, 94, 43), 
                 contrast = 'p', 
                 skew = FALSE, 
                 stratified = TRUE)$estimates |>
  as_tibble()

ratesci::scoreci(x1 = c(4, 21, 11), 
                 n1 = c(17, 94, 43), 
                 wt = c(1,1,1), 
                 contrast = 'p', 
                 skew = FALSE, 
                 stratified = TRUE)$estimates |>
  as_tibble()
```

# References
