---
title: "Confidence Intervals for Paired Proportions in R"
bibliography: references.bib
csl: nature.csl 
---

# Introduction

This page covers confidence intervals for comparisons of two paired proportions in R, including risk difference (RD) $\theta_{RD} = p_1 - p_2$ or relative risk (RR) $\theta_{RR} = p_1 / p_2$. For odds ratio (OR) in the context of paired data, and of particular interest for case-control studies, the OR is estimated conditional on the number of discordant pairs, calculated as $\theta_{OR} = p_{12} / p_{21}$.

See the [summary page](../method_summary/ci_for_prop_intro.html) for general introductory information on confidence intervals for proportions, including the principles underlying the most common methods.

You may experience paired data in any of the following types of situation:

- Tumour assesssments classified as Progressive Disease or Not Progressive Disease performed by an Investigator and separately by an independent panel.

- A paired case-control study (each subject taking active treatment is matched to a patient taking control)

- A cross-over trial where the same subjects take both medications

In all these cases, the calculated proportions for the 2 groups are not independent, as the outcomes in each pair of observations are expected to be correlated.

Using a cross-over study as our example, a 2 x 2 table can be formed as follows:

+-----------------------+---------------+---------------+---------------+
|                       | Placebo\      | Placebo\      | Total         |
|                       | Response= Yes | Response = No |               |
+=======================+===============+===============+===============+
| Active Response = Yes | r             | s             | r+s           |
+-----------------------+---------------+---------------+---------------+
| Active Response = No  | t             | u             | t+u           |
+-----------------------+---------------+---------------+---------------+
| Total                 | r+t           | s+u           | N = r+s+t+u   |
+-----------------------+---------------+---------------+---------------+

The table below indicates the proportions that are estimated from the data (note the difference in structure compared to the usual 2x2 table for independent proportions).

+-----------------------+---------------+---------------+---------------+
|                       | Placebo\      | Placebo\      | Total         |
|                       | Response= Yes | Response = No |               |
+=======================+===============+===============+===============+
| Active Response = Yes | $p_{11}$      | $p_{12}$      | $p_1$         |
+-----------------------+---------------+---------------+---------------+
| Active Response = No  | $p_{21}$      | $p_{22}$      | $(1-p_1)$     |
+-----------------------+---------------+---------------+---------------+
| Total                 | $p_2$         | $(1-p_2)$     |               |
+-----------------------+---------------+---------------+---------------+

The proportions of subjects responding on each treatment are:

Active: $\hat p_1 = (r+s)/N$ and Placebo: $\hat p_2= (r+t)/N$

The estimated difference between the proportions for each treatment is: $D=\hat p_1 - \hat p_2 = (s-t)/N$

The estimated relative risk is $(r+s)/(r+t)$.

The estimated conditional odds ratio is $s/t$.

# Data used

Worked examples below use the following artificial dataset:

+-----------------------+---------------+---------------+---------------+
|                       | Placebo\      | Placebo\      | Total         |
|                       | Response= Yes | Response = No |               |
+=======================+===============+===============+===============+
| Active Response = Yes | r = 20        | s = 15        | r+s = 35      |
+-----------------------+---------------+---------------+---------------+
| Active Response = No  | t = 6         | u = 5         | t+u = 11      |
+-----------------------+---------------+---------------+---------------+
| Total                 | r+t = 26      | s+u = 20      | N = 46        |
+-----------------------+---------------+---------------+---------------+

The estimated paired RD is (35-26)/46 = 0.196. RR is 35/26 = 1.346, and the conditional odds ratio is 15/6 = 2.5.

# Packages

**The {ratesci} package** includes the `scorepairci()` function for computing approximate Score confidence intervals for paired RD, RR and OR, with or without skewness correction. The `moverpairci()` function provides MOVER methods for paired RD and RR. The convenience functions `rdpairci()`, `rrpairci()` and `orpairci()` may be used for a side-by-side comparison of computed intervals for each contrast, and these include other inferior methods such as Wald and other normal approximations.

**The {contingencytables} package** also provides a selection of different methods, including Wald, MOVER and asymptotic score methods for RD and RR, and transformed Wald, Wilson, Clopper-Pearson and mid-P methods for OR.

**The {PropCIs} package** offers the `diffpropci.Wald.mp` and `diffpropci.mp` functions for RD only, which give the Wald and Agresti-Min adjusted Wald methods respectively.

**The {MKinfer} package** provides `binomDiffCI()` with a `paired` argument to cater for paired data. When the recommended `method = "wilson-cc"` is used, the MOVER-Wilson method for RD is produced, incorporating Newcombe's adjustment to the estimated correlation (note that here the 'cc' does not refer to a 'continuity correction' in the usual sense).

**The {ExactCIdiff} package** produces exact CIs for paired RD with the `PairedCI()` function. However, it should only be used when the sample size is not too large as it can be computationally intensive with prohibitively long run times.

# Paired Proportion Difference

For technical details on the methods for calculating confidence intervals for a paired RD see the corresponding [SAS page](https://psiaims.github.io/CAMIS/SAS/ci_for_prop.html#methods-for-calculating-confidence-intervals-for-a-matched-pair-proportion).

From our example dataset, we have estimates as follows: $\hat p_1 = (r+s)/n$ = 35/46 = 0.761 and $\hat p_2= (r+t)/n$ = 26/46 = 0.565

Difference = 0.761-0.565 = 0.196, then selected confidence intervals can be obtained from R as shown below:

```{r}
#| echo: false
#| include: false
library(tidyverse)
library(ratesci)
library(contingencytables)
options(digits=6)
```

### Example code for {ratesci}

```{r}

x <- c(20, 15, 6, 5)

# Selected methods for proximate coverage:
ratesci::rdpairci(x = x,
                  level = 0.95,
                  precis = 6)

```

### Example code for {contingencytables}

```{r}
xmat <- matrix(x, nrow = 2, byrow = TRUE)
contingencytables::the_paired_2x2_table_CIs_difference(xmat)
```

# Paired Relative Risk

For technical details on the methods for calculating confidence intervals for a paired RR see the corresponding [SAS page](https://psiaims.github.io/CAMIS/SAS/ci_for_prop.html#methods-for-calculating-confidence-intervals-for-a-matched-pair-proportion).

### Example code for {ratesci}

```{r}

# Selected methods for proximate coverage:
ratesci::rrpairci(x = x,
                  level = 0.95,
                  precis = 6)
```

### Example code for {contingencytables}

```{r}

contingencytables::the_paired_2x2_table_CIs_ratio(xmat)

```

# Conditional Odds Ratio

For technical details on the methods for calculating confidence intervals for a conditional OR from paired date see the corresponding [SAS page](https://psiaims.github.io/CAMIS/SAS/ci_for_prop.html#methods-for-calculating-confidence-intervals-for-a-matched-pair-proportion).

### Example code for {ratesci}

```{r}

# Selected methods for proximate coverage:
ratesci::orpairci(x = x,
                  level = 0.95,
                  precis = 6)

```

### Example code for {contingencytables}

```{r}
contingencytables::the_paired_2x2_table_CIs_OR(xmat)
```

# Continuity Adjusted Methods

There are relatively few methods widely available for aligning the minimum coverage with the nominal confidence level. The most versatile option is to use functions from the {ratesci} package, which provides optional continuity adjustments, on a sliding scale from 0 to $0.5/N$, for any of the Asymptotic Score or MOVER methods for any contrast.

### Example code for {ratesci}

```{r}
# Selected methods for conservative coverage 

# Using the conventional 0.5 
ratesci::rdpairci(x = x,
                  level = 0.95,
                  cc =0.5,
                  precis = 6)

```

```{r}
# Using an intermediate adjustment of magnitude 0.25 

ratesci::rdpairci(x = x,
                  level = 0.95,
                  cc =0.25,
                  precis = 6)

```

# Consistency with Hypothesis Tests

## Test for association

The Asymptotic Score methods for all contrasts based on paired data are inherently consistent with the McNemar $\chi^2$ test if the 'N-1' adjustment is omitted with `bcf` = FALSE.

## Non-inferiority test

One important use for CIs for proportions is in the analysis of clinical trials aiming to demonstrate non-inferiority. The Asymptotic Score methods are naturally suited for this purpose, as they are derived by inverting a score test statistic, so that the hypothesis test for any null hypothesis value of the contrast parameter is guaranteed to be consistent with the confidence interval. A one-sided non-inferiority test may be obtained with the `scorepairci` function (for example, with a non-inferiority margin of RD = 0.4) as follows:

```{r}
ratesci::scorepairci(x = x,
                  level = 0.95,
                  theta0 = 0.4,
                  precis = 6)$pval

```

# References
