---
title: "Correlation Analysis Using R"
---

```{r}
#| echo: FALSE
#| include: FALSE
library(tidyverse)
```

The most commonly used correlation analysis methods in clinical trials include:

-   Pearson correlation coefficient: product moment coefficient between two continuous variables, measuring linear associations.

    \
    $$
    r = \frac{\sum_{i=1}^n (x_i - m_x)(y_i - m_y)}{\sqrt{\sum_{i=1}^n (x_i - m_x)^2\sum_{i=1}^n (y_i - m_y)^2}},$$

    where $x$ and $y$ are observations from two continuous variables of length $n$ and $m_x$ and $m_y$ are their respective means.

-   Spearman correlation coefficient: rank correlation defined through the scaled sum of the squared values of the difference between ranks of two continuous variables.

    $$
    \rho = \frac{\sum_{i=1}^n (x'_i - m_{x'})(y'_i - m_{y'})}{\sqrt{\sum_{i=1}^n (x'_i - m_{x'})^2\sum_{i=1}^n(y'_i - m_{y'})^2}},
    $$

    where $x'$ and $y'$ are the ranks of $x$ and $y$ and $m_{x'}$ and $m_{y'}$ are the mean ranks of $x$ and $y$, respectively.

-   Kendall's rank correlation: rank correlation based on the number of inversions in one ranking as compared with another.

    $$
    \tau = \frac{n_c - n_d}{\frac{1}{2}\,n\,(n-1)},
    $$

    where $n_c$ is the total number of concordant pairs, $n_d$ is the total number of disconcordant pairs and $n$ the total size of observations in $x$ and $y$.

Other association measures are available for count data/contingency tables comparing observed frequencies with those expected under the assumption of independence

-   Fisher exact test

-   Chi-Square statistic

# Example: Lung Cancer Data

*Data source:* *Loprinzi CL. Laurie JA. Wieand HS. Krook JE. Novotny PJ. Kugler JW. Bartel J. Law M. Bateman M. Klatt NE. et al. Prospective evaluation of prognostic variables from patient-completed questionnaires. North Central Cancer Treatment Group. Journal of Clinical Oncology. 12(3):601-7, 1994.*

Survival in patients with advanced lung cancer from the North Central Cancer Treatment Group. Performance scores rate how well the patient can perform usual daily activities.

```{r}
library(survival) 

glimpse(lung)
```

# Overview

`cor()` computes the correlation coefficient between continuous variables `x` and `y`, where `method` chooses which correlation coefficient is to be computed (default: `"pearson"`, `"kendall"`, or `"spearman"`).

`cor.test()` calulates the test for association between paired samples, using one of Pearson's product moment correlation coefficient, Kendall's $\tau$ or Spearman's $\rho$. Besides the correlation coefficient itself, it provides additional information.

Missing values are assumed to be missing completely at random (MCAR). Different strategies are available, see `?cor` for details.

# Pearson Correlation

```{r}
cor.test(x = lung$age, y = lung$meal.cal, method = "pearson") 
```

# Spearman Correlation

```{r}
cor.test(x = lung$age, y = lung$meal.cal, method = "spearman")
```

Note: Exact p-values require unanimous ranks.

# Kendall's rank correlation

```{r}
cor.test(x = lung$age, y = lung$meal.cal, method = "kendall")
```

# Interpretation of correlation coefficients

Correlation coefficient is comprised between -1 and 1:

-   $-1$ indicates a strong negative correlation

-   $0$ means that there is no association between the two variables

-   $1$ indicates a strong positive correlation

# 
