---
title: "Correlation Analysis Using R"
---

```{r}
#| echo: FALSE
#| include: FALSE

library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

The most commonly used correlation analysis methods in clinical trials include:

-   Pearson correlation coefficient: product moment coefficient between two continuous variables, measuring linear associations

-   Spearman correlation coefficient: rank correlation defined through the scaled sum of the squared values of the difference between ranks of two continuous variables

-   Kendall's rank correlation: rank correlation based on the number of inversions in one ranking as compared with another

-   Chi-Square statistic: association measure for categorical variables comparing observed frequencies with those expected under the assumption of independence

-   C-index: concordance coefficient defined as the proportion of concordant pairs divided by the total number of possible evaluation pairs.

# Example Data

Survival in patients with advanced lung cancer from the North Central Cancer Treatment Group. Performance scores rate how well the patient can perform usual daily activities.

Data source: Loprinzi CL. Laurie JA. Wieand HS. Krook JE. Novotny PJ. Kugler JW. Bartel J. Law M. Bateman M. Klatt NE. et al. Prospective evaluation of prognostic variables from patient-completed questionnaires. North Central Cancer Treatment Group. Journal of Clinical Oncology. 12(3):601-7, 1994.

```{r}
library(survival) 

glimpse(lung)
```

# Correlation Analysis of Continuous Data

## Pearson Correlation

```{r}
cor.test(x = lung$age, y = lung$meal.cal, method = "pearson") 
```

## Spearman Correlation

```{r}
cor.test(x = lung$age, y = lung$meal.cal, method = "spearman")
```

Note: Exact p-values require unanimous ranks.

## Interpretation of correlation coefficients

Correlation coefficient is comprised between -1 and 1:

-   $-1$ indicates a strong negative correlation

-   $0$ means that there is no association between the two variables

-   $1$ indicates a strong positive correlation
