---
title: "Regression for Count Data"
---

```{r}
#| echo: FALSE
#| include: FALSE
library(tidyverse)
```

The most commonly used models for count data in clinical trials include:

-   Poisson regression: assumes the response variable $Y$ has a Poisson distribution, which is linked using the logarithm with explanatory variables $\bf{x}$.

$$
\text{log}(E(Y|x))= \beta_0 + \beta' x, \; i = 1,\ldots,n 
$$

-   Quasi-Poisson regression: Poisson model that allows overdispersion, i.e. dispersion parameter is not fixed at one.
-   Negative-Binomial regression: popular generalization which loosens the assumption that the variance is equal to the mean made by the Poisson model.

Other models include hurdle or zero-inflated models, if data have more zero observations than expected.

# Example: Familial Andenomatous Polyposis Data

*Data source: F. M. Giardiello, S. R. Hamilton, A. J. Krush, S. Piantadosi, L. M. Hylind, P. Celano, S. V. Booker, C. R. Robinson and G. J. A. Offerhaus (1993), Treatment of colonic and rectal adenomas with sulindac in familial adenomatous polyposis. New England Journal of Medicine, 328(18), 1313--1316.*

Data from a placebo-controlled trial of a non-steroidal anti-inflammatory drug in the treatment of familial andenomatous polyposis (FAP). (see `?polyps` for details).

```{r}
data("polyps", package="HSAUR2")
glimpse(polyps)
```

We analyze the number of colonic polyps at 12 months in dependency of treatment and age of the patient.

```{r}
polyps %>% 
  ggplot(aes(y = number, x = age, color = treat)) + 
  geom_point() + theme_minimal()
```

# Model Fit

We fit a generalized linear model for `number` using the Poisson distribution with default log link.

```{r}
# Poisson
m1 <- glm(number ~ treat + age, data = polyps, family = poisson)
summary(m1)
```

The parameter estimates are on log-scale. For better interpretation, we can exponentiate these estimates, to obtain estimates and provide $95$% confidence intervals:

```{r}
# OR and CI
exp(coef(m1))
exp(confint(m1))
```

Predictions for number of colonic polyps given a new 25-year-old patient on either treatment using `predict()`:

```{r}
# new 25 year old patient
new_pt <- data.frame(treat = c("drug","placebo"), age=25)
predict(m1, new_pt, type = "response")
```

# Modelling Overdispersion

Poisson model assumes that mean and variance are equal, which can be a very restrictive assumption. One option to relax the assumption is adding a overdispersion constant to the relationship, i.e. $\text{Var}(\text{response}) = \phi\cdot \mu$, which results in a quasipoisson model:

```{r}
# Quasi poisson
m2 <- glm(number ~ treat + age, data = polyps, family = quasipoisson)
summary(m2)
```

Alternatively, we can explicitly model the count data with overdispersion using the negative Binomial model. In this case, the overdispersion is a function of both $\mu$ and $\mu^2$:

$$
\text{Var}(\text{response}) = \mu + \kappa\,\mu^2.
$$

```{r}
# Negative Binomial
m3 <- MASS::glm.nb(number ~ treat + age, data = polyps)
summary(m3)
```

Both model result very similar parameter estimates, but vary in estimates for their respective standard deviation.
