---
title: "Regression for Count Data"
---

```{r}
#| echo: false
#| include: false
library(tidyverse)
library(MASS)
library(HSAUR2)
```

The most commonly used models for count data in clinical trials include:

-   Poisson regression: assumes the response variable $Y$ has a Poisson distribution, which is linked using the logarithm with explanatory variables $\bf{x}$.

$$
\text{log}(E(Y|x))= \beta_0 + \beta' x, \; i = 1,\ldots,n 
$$

-   Quasi-Poisson regression: Poisson model that allows overdispersion, i.e. dispersion parameter is not fixed at one.
-   Negative-Binomial regression: popular generalization which loosens the assumption that the variance is equal to the mean made by the Poisson model.

Other models include hurdle or zero-inflated models, if data have more zero observations than expected.

# Example: Familial Andenomatous Polyposis Data

*Data source: F. M. Giardiello, S. R. Hamilton, A. J. Krush, S. Piantadosi, L. M. Hylind, P. Celano, S. V. Booker, C. R. Robinson and G. J. A. Offerhaus (1993), Treatment of colonic and rectal adenomas with sulindac in familial adenomatous polyposis. New England Journal of Medicine, 328(18), 1313--1316.*

Data from a placebo-controlled trial of a non-steroidal anti-inflammatory drug in the treatment of familial andenomatous polyposis (FAP). (see `?polyps` for details).

```{r}
polyps <- HSAUR2::polyps
glimpse(polyps)
```

We analyze the number of colonic polyps at 12 months in dependency of treatment and age of the patient.

```{r}
polyps |>
  ggplot(aes(y = number, x = age, color = treat)) +
  geom_point() +
  theme_minimal()
```

# Poisson Distribution Models

We fit a generalized linear model for `number` using the Poisson distribution with default log link.

```{r}
# Poisson
m1 <- stats::glm(number ~ treat + age, data = polyps, family = poisson)
summary(m1)
```

The parameter estimates are on log-scale. For better interpretation, we can exponentiate these estimates, to obtain estimates and provide $95$% confidence intervals:

```{r}
# OR and CI
exp(coef(m1))
exp(confint(m1))
```

Predictions for number of colonic polyps given a new 25-year-old patient on either treatment using `predict()`:

```{r}
# new 25 year old patient
new_pt <- data.frame(treat = c("drug", "placebo"), age = 25)
predict(m1, new_pt, type = "response")
```

# Modelling Overdispersion

## Quasi Poisson Modelling

Poisson model assumes that mean and variance are equal, which can be a very restrictive assumption. One option to relax the assumption is adding a overdispersion constant to the relationship, i.e. $\text{Var}(\text{response}) = \phi\cdot \mu$, which results in a quasipoisson model:

```{r}
# Quasi poisson
m2 <- stats::glm(number ~ treat + age, data = polyps, family = quasipoisson)
summary(m2)
```

## Negative Binomial Modelling

Alternatively, we can explicitly model the count data with overdispersion using the negative Binomial model. In this case, the overdispersion is a function of both $\mu$ and $\mu^2$:

$$
\text{Var}(\text{response}) = \mu + \kappa\,\mu^2.
$$

```{r}
# Negative Binomial
m3 <- MASS::glm.nb(number ~ treat + age, data = polyps)
summary(m3)
```

Both model result very similar parameter estimates, but vary in estimates for their respective standard deviation.

**Note:** the MASS::select() function can cause conflict with the dplyr::select() function. To make your code robust to duplicate functions appearing in different packages, it's good practice to always include the package name in your code prior to the function. For example, use dplyr::select() instead of just select(), and then if you are using MASS() for negative binomial regression then you wont have a conflict with using select() in other parts of your code!

## Negative Binomial Modelling with an offset variable

On clinical trials, subjects are not usually all followed for the full duration of the trial (e.g. for a full 12 months). Instead, it is common to have subjects withdrawing from study or treatment and being impacted by intercurrent events (ICEs) such as rescue medications. This means when we are counting numbers of events, their count may be impacted by the follow up time that we studied them for.

Negative binomial modelling is often adjusted for the follow up time, by including an 'offset' variable as shown below.

```{r}
# Suppose we have a variable with the Follow up period for each subject
# We've just created a pretend one below!
set.seed(12345)
polyps2 <- polyps %>% 
         dplyr::mutate(followup = runif(nrow(.),min=1,max=100)) 

# Negative Binomial with offset of the follow up time per subject
m3 <- MASS::glm.nb(number ~ treat + age + offset(log(followup)), data = polyps2)
summary(m3)
```
