## title: "friedman's test"

### **Getting Started**

Friedman's test is used when you have one within-subjects independent variable with two or more levels and a dependent variable that is not interval and normally distributed (but at least ordinal). To build such unreplicated blocked data, we'll create a data frame called `df_bp` from random number.

```{r}
#| eval: false
#| include: false
library("tidyverse")

```

### Data used {.unnumbered}

`bp` will be the dependent variable. `subjid` serves as block and `time_point` as group.

```{r}
set.seed(123)

df_bp = data.frame(bp = runif(n = 50, 138, 200)) %>% 
  mutate(subjid = as.factor(row_number() %% 5),  
         time_point = as.factor((row_number()-1) %/% 5 + 1))

head(df_bp)
```

### Friedman's test {.unnumbered}

For this example, we're testing for a significant difference in median `bp` of different time points using `friedman_test()` from package *rstatix*.

```{r}
#| eval: false
res.fried <- df_bp %>% 
  rstatix::friedman_test(bp ~ time_point | subjid)
res.fried
```

`friedman_effsize()` can be used to calculated effect size.

```{r}
df_bp %>% rstatix::friedman_effsize(bp ~ time_point | subjid)
```
