---
title: "Group sequential design in R"
editor_options:
  chunk_output_type: console
---

## Group sequential design: time-to-event endpoint

While a group sequential design (GSD) could be applied for different types
of endpoints, here we focus on time-to-event endpoints.

## Available R packages

The commonly used R packages for power and sample size calculations
utilizing a GSD are: [gsDesign](https://keaven.github.io/gsDesign/)
(also has a [web interface](https://rinpharma.shinyapps.io/gsdesign/)),
[gsDesign2](https://merck.github.io/gsDesign2/),
and [rpact](https://www.rpact.org/).

## Design assumptions

Using a toy example, we will assume that a primary objective of a phase III
oncology trial is to compare a new therapy to a control in terms of
progression-free survival (PFS) and overall survival (OS).
Note that, in this example, we have a family of primary endpoints,
i.e., if at least one of the endpoints is successful, the study will be
declared a success. A GSD will be utilized for each endpoint.
PFS will be tested at one interim analysis (IA) for both efficacy and
non-binding futility, while OS will be tested at two IAs for efficacy only.
An O'Brien-Fleming spending function will be used for efficacy testing and a
Hwang-Shih-Decani spending function with $\gamma = -10$ will be used for futility.

Further design assumptions are as follows:

```{r}
# PFS HR = 0.6
hr1_pfs <- 0.6
# Median PFS of 9.4 months in the control arm
med_pfs <- 9.4
# Median follow-up of 10 months for PFS
minfu_pfs <- 10
# Monthly dropout of 0.019 for PFS
do_rate_pfs <- 0.019
# IA timing for PFS is at 75% information fraction
timing_pfs <- c(0.75, 1)
# Power of 95% for PFS
power_pfs <- 0.95

# OS HR = 0.65
hr1_os <- 0.65
# Median OS of 3 years in the control arm
med_os <- 12 * 3
# Median follow-up of 42 months for OS
minfu_os <- 42
# Monthly dropout of 0.001 for OS
do_rate_os <- 0.001
# IA timing for OS is at 60% and 80% information fraction
timing_os <- c(0.6, 0.8, 1)
# Power of 82% for OS
power_os <- 0.82

# Enrollment period of 24 months
enroll_dur <- 24
# 1:1 randomization ratio
rand_ratio <- 1
# alpha level of 1.25% for each endpoint
alphal <- 0.0125
```

We assume that given the above assumptions, we need to calculate the
target number of events for each analysis as well as the total sample size.

## Example code

### Example using gsDesign

- PFS calculations:

```{r}
#| warning=FALSE,
#| message=FALSE
library(gsDesign)
```

```{r}
pfs_gsDesign <- gsSurv(
  k = length(timing_pfs),
  timing = timing_pfs,
  R = enroll_dur,
  eta = do_rate_pfs,
  minfup = minfu_pfs,
  T = enroll_dur + minfu_pfs,
  lambdaC = log(2) / med_pfs,
  hr = hr1_pfs,
  beta = 1 - power_pfs,
  alpha = alphal,
  sfu = sfLDOF,
  sfl = sfHSD,
  sflpar = -10,
  test.type = 4
)

pfs_gsDesign |> gsBoundSummary()
```

- OS calculations:

```{r}
os_gsDesign <- gsSurv(
  k = length(timing_os),
  timing = timing_os,
  R = enroll_dur,
  eta = do_rate_os,
  minfup = minfu_os,
  T = enroll_dur + minfu_os,
  lambdaC = log(2) / med_os,
  hr = hr1_os,
  beta = 1 - power_os,
  alpha = alphal,
  sfu = sfLDOF,
  test.type = 1
)

os_gsDesign |> gsBoundSummary()
```

### Example using gsDesign2

- PFS calculations:

```{r}
#| warning=FALSE,
#| message=FALSE
library(gsDesign2)
library(tibble)
```

```{r}
enroll_rate <- tibble(
  stratum = "All",
  duration = enroll_dur,
  rate = 1
)
fail_rate_pfs <- tibble(
  stratum = "All",
  duration = Inf, # Can be set to `Inf` when proportional hazard is assumed
  fail_rate = log(2) / med_pfs,
  hr = hr1_pfs,
  dropout_rate = do_rate_pfs
)

pfs_gsDesign2 <- gs_design_ahr(
  enroll_rate = enroll_rate,
  fail_rate = fail_rate_pfs,
  ratio = rand_ratio,
  beta = 1 - power_pfs,
  alpha = alphal,
  info_frac = timing_pfs,
  analysis_time = enroll_dur + minfu_pfs,
  upper = gs_spending_bound,
  upar = list(
    sf = gsDesign::sfLDOF,
    total_spend = alphal
  ),
  lower = gs_spending_bound,
  lpar = list(
    sf = gsDesign::sfHSD,
    total_spend = 1 - power_pfs,
    param = -10
  ),
  info_scale = "h0_info"
)

pfs_gsDesign2 |>
  summary() |>
  as_gt()
```

- OS calculations:

```{r}
#| warning=FALSE
fail_rate_os <- tibble(
  stratum = "All",
  duration = Inf, # Can be set to `Inf` when proportional hazard is assumed
  fail_rate = log(2) / med_os,
  hr = hr1_os,
  dropout_rate = do_rate_os
)

os_gsDesign2 <- gs_design_ahr(
  enroll_rate = pfs_gsDesign2$enroll_rate,
  fail_rate = fail_rate_os,
  ratio = rand_ratio,
  beta = 1 - power_os,
  alpha = alphal,
  info_frac = timing_os,
  analysis_time = enroll_dur + minfu_os,
  test_lower = FALSE,
  upper = gs_spending_bound,
  upar = list(
    sf = gsDesign::sfLDOF,
    total_spend = alphal
  ),
  info_scale = "h0_info"
)

os_gsDesign2 |>
  summary() |>
  as_gt()
```

### Example using rpact

- PFS calculations:

```{r}
#| warning=FALSE,
#| message=FALSE
library(rpact)
```

```{r}
pfs_rpact_gsd <- getDesignGroupSequential(
  sided = 1,
  alpha = alphal,
  informationRates = timing_pfs,
  typeOfDesign = "asOF",
  beta = 1 - power_pfs,
  typeBetaSpending = "bsHSD",
  gammaB = -10,
  bindingFutility = FALSE
)

pfs_rpact <- getSampleSizeSurvival(
  design = pfs_rpact_gsd,
  accrualTime = enroll_dur,
  followUpTime = minfu_pfs,
  lambda2 = log(2) / med_pfs,
  hazardRatio = hr1_pfs,
  dropoutRate1 = 0.2,
  dropoutRate2 = 0.2,
  dropoutTime = 12
)

kable(summary(pfs_rpact))
```

Note: the `dropoutRate1`, `dropoutRate2` arguments in `getSampleSizeSurvival()`
refer to the % of drop-outs by the `dropoutTime`, while the `eta` argument
in `gsDesign::gsSurv()` and the `dropout_rate` value in the `fail_rate` argument
in `gsDesign2::gs_design_ahr()` refer to the annual drop-out rate parameter
under the exponential distribution.
In our example, if $X$ is a drop-out time and $X \sim \text{Exponential} (\lambda)$,
we assume that by month 12 the drop-out rate was 20%, which implies:
$P(X\le12) = 1 - e^{-12\lambda} = 0.2 \Rightarrow \lambda = 0.019$.
Due to the above differences, the value $\lambda = 0.019$ was used
in the gsDesign and gsDesign2 example, while 0.2 was used in the rpact example.

- OS calculations:

```{r}
#| warning=FALSE
os_rpact_gsd <- getDesignGroupSequential(
  sided = 1,
  alpha = alphal,
  informationRates = timing_os,
  typeOfDesign = "asOF",
  beta = 1 - power_os
)

os_rpact <- getSampleSizeSurvival(
  design = os_rpact_gsd,
  accrualTime = enroll_dur,
  followUpTime = minfu_os,
  lambda2 = log(2) / med_os,
  hazardRatio = hr1_os,
  dropoutRate1 = 1 - exp(-do_rate_os * 12),
  dropoutRate2 = 1 - exp(-do_rate_os * 12),
  dropoutTime = 12
)

kable(summary(os_rpact))
```
