---
title: "Kruskal Wallis R"
---

## Introduction

The Kruskal-Wallis test is a non-parametric equivalent to the one-way ANOVA. For this example, the data used is a subset of datasets::iris, testing for difference in sepal width between species of flower.

```{r}
#| echo: false
species_n <- c('setosa', 'versicolor', 'virginica')
# fmt: skip
iris_sub <- data.frame(
  Species = c(rep(species_n, each = 6)),
  Sepal_Width = c(
    3.4, 3.0, 3.4, 3.2, 3.5, 3.1, 2.7, 2.9, 2.7, 
    2.6, 2.5, 2.5, 3.0, 3.0, 3.1, 3.8, 2.7, 3.3
  )
)

print(iris_sub)
```

## Implementing Kruskal-Wallis in R

The Kruskal-Wallis test can be implemented in R using stats::kruskal.test. Below, the test is defined using R's formula interface (dependent \~ independent variable) and specifying the data set. The null hypothesis is that the samples are from identical populations.

```{r}
stats::kruskal.test(Sepal_Width ~ Species, data = iris_sub)
```

## Results

As seen above, R outputs the Kruskal-Wallis rank sum statistic (10.922), the degrees of freedom (2), and the p-value of the test (0.004249). Therefore, the difference in population medians is statistically significant at the 5% level.
