---
title: "Logistic Regression in R"
---

```{r}
#| echo: FALSE
#| include: FALSE
library(tidyverse)
```

In binary logistic regression, there is a single binary dependent variable, coded by an indicator variable. For example, if we represent a response as 1 and non-response as 0, then the corresponding probability of response, can be between 0 (certainly not a response) and 1 (certainly a response) - hence the labeling !

The logistic model models the log-odds of an event as a linear combination of one or more independent variables (explanatory variables). If we observed $(y_i, x_i),$ where $y_i$ is a Bernoulli variable and $x_i$ a vector of explanatory variables, the model for $\pi_i = P(y_i=1)$ is

$$
\text{logit}(\pi_i)= \log\left\{ \frac{\pi_i}{1-\pi_i}\right\} = \beta_0 + \beta x_i, i = 1,\ldots,n 
$$

The model is especially useful in case-control studies and leads to the effect of risk factors by odds ratios.

# Example: Lung Cancer Data

*Data source: Loprinzi CL. Laurie JA. Wieand HS. Krook JE. Novotny PJ. Kugler JW. Bartel J. Law M. Bateman M. Klatt NE. et al. Prospective evaluation of prognostic variables from patient-completed questionnaires. North Central Cancer Treatment Group. Journal of Clinical Oncology. 12(3):601-7, 1994.*

Survival in patients with advanced lung cancer from the North Central Cancer Treatment Group. Performance scores rate how well the patient can perform usual daily activities (see `?lung` for details).

```{r}
library(survival) 
glimpse(lung)
```

# Model Fit

We analyze the event of weight gain (or staying the same weight) in lung cancer patients in dependency of age, sex, ECOG performance score and calories consumed at meals. In the original data, a positive number for the `wt.loss` variable is a weight loss, negative number is a gain. We start by dichotomising the response such that a result \>0 is a weight loss, \<= weight gain and creating a factor variable `wt_grp`.

One of the most important things to remember is to ensure you tell R what your event is ! We want to model Events / Non-events, and hence your reference category for `wt_grp` dichotomous variable below is the weight loss level. Therefore, by telling R that your reference category is weight loss, you are effectively telling R that your Event = Weight Gain !

```{r}
lung2 <- survival::lung %>% 
  mutate(
    wt_grp = factor(wt.loss > 0, labels = c("weight loss", "weight gain"))
  ) 

#specify that weight loss should be used as baseline level (i.e we want to model weight gain as the event)
lung2$wt_grp <- relevel(lung2$wt_grp, ref='weight loss')

m1 <- glm(wt_grp ~ age + sex + ph.ecog + meal.cal, data = lung2, family = binomial(link="logit"))
summary(m1)
```

The model summary contains the parameter estimates $\beta_j$ for each explanatory variable $x_j$, corresponding to the log-odds for the response variable to take the value $1$, conditional on all other explanatory variables remaining constant. For better interpretation, we can exponentiate these estimates, to obtain estimates for the odds instead and provide $95$% confidence intervals:

```{r}
exp(coef(m1))
exp(confint(m1))
```

# Model Comparison

To compare two logistic models, the `residual deviances` (-2 \* log likelihoods) are compared against a $\chi^2$-distribution with degrees of freedom calculated using the difference in the two models' parameters. Below, the only difference is the inclusion/exclusion of age in the model, hence we test using $\chi^2$ with 1 df. Here testing at the $5$% level.

```{r}
m2 <- glm(wt_grp ~ sex + ph.ecog + meal.cal, data = lung2, family = binomial(link="logit"))
summary(m2)

anova(m1, m2, test = "LRT")
```

Stackexchange [here](https://stats.stackexchange.com/questions/59879/logistic-regression-anova-chi-square-test-vs-significance-of-coefficients-ano) has a good article describing this method and the difference between comparing 2 models using the likelihood ratio tests versus using wald tests and Pr\>chisq (from the maximum likelihood estimate). Note: `anova(m1, m2, test = "Chisq")` and using `test="LRT"` as above are synonymous in this context.

# Prediction

Predictions from the model for the log-odds of a patient with new data to experience a weight loss are derived using `predict()`:

```{r}
# new female, symptomatic but completely ambulatory patient consuming 2500 calories
new_pt <- data.frame(sex=2, ph.ecog=1, meal.cal=2500)
predict(m2, new_pt, type = "response")
```
