---
title: "Logistic Regression"
---

```{r}
#| echo: FALSE
#| include: FALSE
library(tidyverse)
```

A model of the dependence of binary variables on explanatory variables. The logit of expectation is explained as linear for of explanatory variables. If we observed $(y_i, x_i),$ where $y_i$ is a Bernoulli variable and $x_i$ a vector of explanatory variables, the model for $\pi_i = P(y_i=1)$ is

$$
\text{logit}(\pi_i)= \log\left\{ \frac{\pi_i}{1-\pi_i}\right\} = \beta_0 + \beta x_i, i = 1,\ldots,n 
$$ 

The model is especially useful in case-control studies and leads to the effect of risk factors by odds ratios.

# Example: Lung Cancer Data

*Data source: Loprinzi CL. Laurie JA. Wieand HS. Krook JE. Novotny PJ. Kugler JW. Bartel J. Law M. Bateman M. Klatt NE. et al. Prospective evaluation of prognostic variables from patient-completed questionnaires. North Central Cancer Treatment Group. Journal of Clinical Oncology. 12(3):601-7, 1994.*

Survival in patients with advanced lung cancer from the North Central Cancer Treatment Group. Performance scores rate how well the patient can perform usual daily activities (see `?lung` for details).

```{r}
library(survival) 
glimpse(lung)
```

# Model Fit

We analyze the weight loss in lung cancer patients in dependency of age, sex, ECOG performance score and calories consumed at meals.

```{r}
lung2 <- survival::lung %>% 
  mutate(
    wt_grp = factor(wt.loss > 0, labels = c("weight loss", "weight gain"))
  ) 


m1 <- glm(wt_grp ~ age + sex + ph.ecog + meal.cal, data = lung2, family = binomial(link="logit"))
summary(m1)
```

The model summary contains the parameter estimates $\beta_j$ for each explanatory variable $x_j$, corresponding to the log-odds for the response variable to take the value $1$, conditional on all other explanatory variables remaining constant. For better interpretation, we can exponentiate these estimates, to obtain estimates for the odds instead and provide $95$% confidence intervals:

```{r}
exp(coef(m1))
exp(confint(m1))
```

# Model Comparison

To compare two logistic models, one tests the difference in residual variances from both models using a $\chi^2$-distribution with a single degree of freedom (here at the $5$% level):

```{r}
m2 <- glm(wt_grp ~ sex + ph.ecog + meal.cal, data = lung2, family = binomial(link="logit"))
summary(m2)

anova(m1, m2, test = "Chisq")
```

# Prediction

Predictions from the model for the log-odds of a patient with new data to experience a weight loss are derived using `predict()`:

```{r}
# new female, symptomatic but completely ambulatory patient consuming 2500 calories
new_pt <- data.frame(sex=2, ph.ecog=1, meal.cal=2500)
predict(m2, new_pt, type = "response")
```
