---
title: "Marginal Homogeneity Tests"
---

This page is solely based on **coin** package documentation including data samples which are generated inline.

`coin::mh_test()` provides the McNemar test, the Cochran Q test, the Stuart(-Maxwell) test and the Madansky test of interchangeability. A general description of these methods is given by Agresti (2002).

The null hypothesis of marginal homogeneity is tested. If formula interface is used, the response variable and the measurement conditions are given by `y` and `x`, respectively, and `block` is a factor where each level corresponds to exactly one subject with repeated measurements: `coin::mh_test(y ~ x | block, data, subset = NULL, ...)`. We can also directly pass an object of class `"table"`.

`coin::mh_test()` computes different tests depending on `x` and `y`:

-   McNemar test (McNemar, 1947) when both `y` and `x` are binary factors;

-   Cochran Q test (Cochran, 1950) when `y` is a binary factor and `x` is a factor with an arbitrary number of levels;

-   Stuart-Maxwell test (Stuart, 1955; Maxwell, 1970) when `y` is a factor with an arbitrary number of levels and `x` is a binary factor;

-   Madansky test of interchangeability (Madansky, 1963), which implies marginal homogeneity, when both `y` and `x` are factors with an arbitrary number of levels.

The conditional null distribution of the test statistic is used to obtain p-values and an asymptotic approximation of the exact distribution is used by default (`distribution = "asymptotic"`). Alternatively, the distribution can be approximated via Monte Carlo resampling or computed exactly for univariate two-sample problems (McNemar test) by setting distribution to `"approximate"` or `"exact"`, respectively.

## McNemar test

McNemar test is calculated [without continuity correction]{.underline} which corresponds to SAS FREQ procedure (see [R v SAS McNemar’s test](https://psiaims.github.io/CAMIS/Comp/r-sas_mcnemar.html) for more details).

```{r}
## Performance of prime minister
## Agresti (2002, p. 409)
performance <- matrix(
  c(794, 150,
    86, 570),
  nrow = 2, 
  byrow = TRUE,
  dimnames = list(
    "First" = c("Approve", "Disprove"),
    "Second" = c("Approve", "Disprove")
  )
)
performance <- as.table(performance)

performance

## Asymptotic McNemar Test
# Corresponds to SAS FREQ procedure
coin::mh_test(performance)

## Approximative McNemar Test
coin::mh_test(
  performance, 
  distribution = coin::approximate(nresample = 10000)
)

## Exact McNemar Test
coin::mh_test(performance, distribution = "exact")
```

## Cochran Q test

```{r}
## Effectiveness of different media for the growth of diphtheria
## Cochran (1950, Tab. 2)
cases <- c(4, 2, 3, 1, 59)
n <- sum(cases)
cochran <- data.frame(
  diphtheria = factor(
    unlist(rep(list(c(1, 1, 1, 1),
                    c(1, 1, 0, 1),
                    c(0, 1, 1, 1),
                    c(0, 1, 0, 1),
                    c(0, 0, 0, 0)),
                cases)
    )
  ),
  media = factor(rep(LETTERS[1:4], n)),
  case =  factor(rep(seq_len(n), each = 4))
)

head(cochran)

## Asymptotic Cochran Q test (Cochran, 1950, p. 260)
coin::mh_test(
  diphtheria ~ media | case, 
  data = cochran
) 

## Approximative Cochran Q test
mt <- coin::mh_test(
  diphtheria ~ media | case, 
  data = cochran,
  distribution = coin::approximate(nresample = 10000)
)
coin::pvalue(mt)             # standard p-value
coin::midpvalue(mt)          # mid-p-value
coin::pvalue_interval(mt)    # p-value interval
coin::size(mt, alpha = 0.05) # test size at alpha = 0.05 using the p-value
```

## Stuart-Maxwell test

```{r}
## Opinions on Pre- and Extramarital Sex
## Agresti (2002, p. 421)
opinions <- c("Always wrong", "Almost always wrong",
              "Wrong only sometimes", "Not wrong at all")
PreExSex <- matrix(
  c(144, 33, 84, 126,
    2,  4, 14,  29,
    0,  2,  6,  25,
    0,  0,  1,   5),
  nrow = 4,
  dimnames = list(
    "Premarital Sex" = opinions,
    "Extramarital Sex" = opinions
  )
)
PreExSex <- as.table(PreExSex)

PreExSex

## Asymptotic Stuart test
coin::mh_test(PreExSex)

## Asymptotic Stuart-Birch test
## Note: response as ordinal
coin::mh_test(
  PreExSex, 
  scores = list(response = 1:length(opinions))
)
```

## Madansky test of interchangeability

```{r}
## Vote intention
## Madansky (1963, pp. 107-108)
vote <- array(
    c(120, 1,  8, 2,   2,  1, 2, 1,  7,
        6, 2,  1, 1, 103,  5, 1, 4,  8,
       20, 3, 31, 1,   6, 30, 2, 1, 81),
    dim = c(3, 3, 3),
    dimnames = list(
          "July" = c("Republican", "Democratic", "Uncertain"),
        "August" = c("Republican", "Democratic", "Uncertain"),
          "June" = c("Republican", "Democratic", "Uncertain")
    )
)
vote <- as.table(vote)

vote

## Asymptotic Madansky test (Q = 70.77)
coin::mh_test(vote)

## Cross-over study
## http://www.nesug.org/proceedings/nesug00/st/st9005.pdf (link is dead now)
dysmenorrhea <- array(
  c(6, 2, 1,  3, 1, 0,  1, 2, 1,
    4, 3, 0, 13, 3, 0,  8, 1, 1,
    5, 2, 2, 10, 1, 0, 14, 2, 0),
  dim = c(3, 3, 3),
  dimnames =  list(
    "Placebo" = c("None", "Moderate", "Complete"),
    "Low dose" = c("None", "Moderate", "Complete"),
    "High dose" = c("None", "Moderate", "Complete")
  )
)
dysmenorrhea <- as.table(dysmenorrhea)

dysmenorrhea

## Asymptotic Madansky-Birch test (Q = 53.76)
## Note: response as ordinal
coin::mh_test(
  dysmenorrhea, 
  scores = list(response = 1:3)
)

## Asymptotic Madansky-Birch test (Q = 47.29)
## Note: response and measurement conditions as ordinal
coin::mh_test(
  dysmenorrhea, 
  scores = list(response = 1:3, conditions = 1:3)
)
```

## Reference

Hothorn T, Hornik K, van de Wiel MA, Zeileis A (2006). A Lego system for conditional inference. The American Statistician, 60 (3), 257-263. doi:10.1198/000313006X118430 <https://doi.org/10.1198/000313006X118430>

Agresti, A. (2002). Categorical Data Analysis, Second Edition. Hoboken, New Jersey: John Wiley & Sons.

Birch, M. W. (1965). The detection of partial association, II: The general case. Journal of the Royal Statistical Society B 27(1), 111–124. doi:10.1111/j.2517-6161.1965.tb00593.x

Cochran, W. G. (1950). The comparison of percentages in matched samples. Biometrika 37(3/4), 256–266. doi:10.1093/biomet/37.3-4.256

Madansky, A. (1963). Tests of homogeneity for correlated samples. Journal of the American Statistical Association 58(301), 97–119. doi:10.1080/01621459.1963.10500835

Maxwell, A. E. (1970). Comparing the classification of subjects by two independent judges. British Journal of Psychiatry 116(535), 651–655. doi:10.1192/bjp.116.535.651

McNemar, Q. (1947). Note on the sampling error of the difference between correlated proportions or percentages. Psychometrika 12(2), 153–157. doi:10.1007/BF02295996

Stuart, A. (1955). A test for homogeneity of the marginal distributions in a two-way classification. Biometrika 42(3/4), 412–416. doi:10.1093/biomet/42.3-4.412

White, A. A., Landis, J. R. and Cooper, M. M. (1982). A note on the equivalence of several marginal homogeneity test criteria for categorical data. International Statistical Review 50(1), 27–34. doi:10.2307/1402457

::: {.callout-note collapse="true" title="Session Info"}
```{r}
#| echo: false

# List all the packages needed 
si <- sessioninfo::session_info(c(
  "coin"
))
si
```
:::
