---
title: "Multiple Imputaton: Linear Regression"
---

## Overview

Multiple imputation with regression is one step further from mean imputation (i.e. by a single value: the average of observed). In the case for continuous, normally distributed variable, linear regression can use information from other variables hence *could* be closer to the true missing values.


## Imputation with `mice`

[mice](https://amices.org/mice/index.html) is a powerful R package developed by Stef van Buuren, Karin Groothuis-Oudshoorn and other contributors. **Regression methods** (continuous, normal outcome) are implemented in `mice` with methods starting with `norm`.


* Linear regression without parameter uncertainty, [mice.impute.norm.nob](https://amices.org/mice/reference/mice.impute.norm.nob.html)
* Linear regression through prediction, [mice.impute.norm.predict](https://amices.org/mice/reference/mice.impute.norm.predict.html)
* Bayesian linear regression, [mice.impute.norm](https://amices.org/mice/reference/mice.impute.norm.html)
* Linear regression bootstrap, [mice.impute.norm.boot](https://amices.org/mice/reference/mice.impute.norm.boot.html)


## Example

Here I use the small dataset `nhanes` included in `mice` package. It has 25 rows, and three out of four variables have missings. 

The original NHANES data is a large national level survey, some are publicly available via R package `nhanes`.

```{r}
#| label: data
#| echo: true
#| eval: true

library(mice)

# load example dataset from mice
head(nhanes)
summary(nhanes)
```

Examine missing pattern with `md.pattern(data)`.


```{r}
#| label: missing-pattern
#| echo: true
#| eval: true

# 27 missing in total
# by col: 8 for hyp, 9 for bmi, 10 for chl
# by row: n missing numbers

md.pattern(nhanes)
```


### Regression without parameter uncertainty

We can generate two imputed datasets by setting `m=2`. 

There is a certain level of randomness, so would be a good idea to set seed.

```{r}
#| label: impute-reg-1
#| echo: true
#| eval: true

set.seed(1)
impr0 <- mice(nhanes, method = 'norm.nob', m=2, maxit = 1)
impr0

nhanes_impr0 <- complete(impr0) # by default, returns the first imputation
nhanes_impr0
```
When we have two imputed datasets, we can check the values for each of the variables. For example, extract `bmi` variable from the imputed data `imp`,

```{r}
#| label: impute-reg-2
#| echo: true
#| eval: true

# two imputed datasets (m=2)
impr0$imp$bmi
```

We can also specify which imputed dataset to use as our complete data. Set index to 0 (`action = 0`) returns the original dataset with missing values.

Here we check which of the imputed data is being used as the completed dataset. First take a note of the row IDs (based on `bmi`, for example). Then we generate completed dataset. 

* if no `action` argument is set, then it returns the first imputation by default
* `action=0` corresponds to the original data with missing values

```{r}
#| label: impute-reg-3
#| echo: true
#| eval: true

# check which imputed data is used for the final result, take note of row id
id_missing <- which(is.na(nhanes$bmi))
id_missing

nhanes_impr0_action0 <- complete(impr0, action = 0) 
nhanes_impr0_action0[id_missing, ] # original data with missing bmi

nhanes_impr0_action1 <- complete(impr0, action = 1) 
nhanes_impr0_action1[id_missing, ] # using first imputation

nhanes_impr0_action2 <- complete(impr0, action = 2) 
nhanes_impr0_action2[id_missing, ] # using second imputation
```


### Other imputation by linear regression

Other various of imputaton via linear regression can be implemented simply by changing the method argument.

* Linear regression through prediction, [mice.impute.norm.predict](https://amices.org/mice/reference/mice.impute.norm.predict.html)
* Bayesian linear regression, [mice.impute.norm](https://amices.org/mice/reference/mice.impute.norm.html)
* Linear regression bootstrap, [mice.impute.norm.boot](https://amices.org/mice/reference/mice.impute.norm.boot.html)

```{r}
#| label: impute-reg-pred
#| echo: true
#| eval: true
impr <- mice(nhanes, method = 'norm.predict', m=1, maxit=1)
impr$imp$bmi
```

Bayesian linear regression

```{r}
#| label: impute-reg-bayes
#| echo: true
#| eval: true
impb <- mice(nhanes, method = 'norm', m=1, maxit=1)
impb$imp$bmi
# nhanes_impb <- complete(impb)
```

Bootstrap

```{r}
#| label: impute-reg-bt
#| echo: true
#| eval: true

impbt <- mice(nhanes, method = 'norm.boot', m=1, maxit=1)
impbt$imp$bmi
```






# Reference

Stef van Buuren, Karin Groothuis-Oudshoorn (2011). mice: Multivariate Imputation by Chained Equations in R. Journal of Statistical Software, 45(3), 1-67. DOI 10.18637/jss.v045.i03

