---
title: "MMRM in R"
---

### Fitting the MMRM in R

#### Using the `nlme::gls` function

The code below implements an MMRM fit in R with the `nlme::gls` function.

```{r}
#| eval: false
#| echo: true
gls(model = CHG ~ TRTP + AVISITN + TRTP:AVISITN + AVISITN + BASE,
    data = data,
    correlation = corSymm(form = ~1|SUBJID),
    weights = varIdent(form = ~1|AVISITN),
    control = glsControl(opt = "optim"),
    method = "REML",
    na.action = "na.omit")
```

here we can swap out `corSymm` for `corCompSymm` to give the compound symmetry structure or `corCAR1` for autoregressive of first order (AR(1)).

#### Using the `lme4::lmer` function

An alternative way to fit an MMRM with unstructured covariance matrices is to use the `lme4::lmer` function as described by Daniel Sabanes Bove in his R in Pharma talk from 2020 see [here](https://github.com/rinpharma/2020_presentations/blob/main/talks_folder/2020-Sabanes_Bove-Implementing_MMRM_in_R.pdf). The relevance of this fit is apparent when we consider the availability of the Kenward-Roger's degrees of freedom for the MMRM in R, which at the time of writing, were not yet available for the `nlme::gls` function via the `pbkrtest` package (see [here](https://people.math.aau.dk/~sorenh/software/pbkrtest/#NEWS)).

```{r}
#| eval: false
#| echo: true
lmer(CHG ~ TRTA * VISIT + VISIT + BASE + (0 + VISIT|SUBJID),
     data = data,
     control = lmerControl(check.nobs.vs.nRE = "ignore"),
     na.action = na.omit)
```

#### Extracting effect estimates using `emmeans`

In order to extract relevant marginal means (LSmeans) and contrasts we can use the `emmeans` package. Below we start by constructing a `ref_grid` used to make explicit just how the predictions are generated across the levels of `TRTP` and `AVISITN`. The `emmeans` function permits various marginal means to be extracted depending on the `formula` provided and the following `pairs()` function call derives relevant contrasts. Note that more control can be obtained by calling the `contrast()` function.

```{r}
#| eval: false
#| echo: true
mod_grid <- ref_grid(model, data = data, mode = "df.error")
mod_emm <- emmeans(mod_grid, ~TRTP * AVISITN, mode = "df.error") 
pairs(mod_emm) 
```
