---
title: "MMRM in R"
---

### Fitting the MMRM in R
Longitudinal data are frequently collected as part of clinical trials giving multiple visit records per subject in a study with a serial trace of some outcome of interest. The Mixed Model for Repeated Measures (MMRM) model is ubiquitous in this context and has a very standard implementation in SAS using `proc mixed`.

#### Using the `lme4::lmer` function

An alternative way to fit an MMRM with unstructured covariance matrices is to use the `lme4::lmer` function as described by Daniel Sabanes Bove in his R in Pharma talk from 2020 see [here](https://github.com/rinpharma/2020_presentations/blob/main/talks_folder/2020-Sabanes_Bove-Implementing_MMRM_in_R.pdf). The relevance of this fit is apparent when we consider the availability of the Kenward-Roger's degrees of freedom for the MMRM in R, which at the time of writing, were not yet available for the `nlme::gls` function via the `pbkrtest` package (see [here](https://people.math.aau.dk/~sorenh/software/pbkrtest/#NEWS)).

```{r}
#| eval: false
#| echo: true
lmer(CHG ~ TRTA * VISIT + VISIT + BASE + (0 + VISIT|SUBJID),
     data = data,
     control = lmerControl(check.nobs.vs.nRE = "ignore"),
     na.action = na.omit)
```

#### Extracting effect estimates using `emmeans`

In order to extract relevant marginal means (LSmeans) and contrasts we can use the `emmeans` package. Below we start by constructing a `ref_grid` used to make explicit just how the predictions are generated across the levels of `TRTP` and `AVISITN`. The `emmeans` function permits various marginal means to be extracted depending on the `formula` provided and the following `pairs()` function call derives relevant contrasts. Note that more control can be obtained by calling the `contrast()` function.

```{r}
#| eval: false
#| echo: true
mod_grid <- ref_grid(model, data = data, mode = "df.error")
mod_emm <- emmeans(mod_grid, ~TRTP * AVISITN, mode = "df.error") 
pairs(mod_emm) 
```

### Using the `mmrm` package

The code below implement an MMRM fit in R using the {mmrm} package. The {mmrmm} package is developed as part of an ASA working group. For a full detailed description of the {mmrm} package please see the [documentation](https://openpharma.github.io/mmrm/latest-tag/) and [package code repository](https://github.com/openpharma/mmrm).

```{r}
#| eval: false
#| echo: true

library(mmrm)
fit <- mmrm(
  formula = FEV1 ~ RACE + SEX + ARMCD * AVISIT + us(AVISIT | USUBJID),
  data = fev_data
)
```
