---
title: "Sample Size Calculation for Average Bioequivalence"
---

## Regulatory Requirements

The most unambiguous requirements are mentioned in [FDA Guidance for Industry. Statistical Approaches to Establishing Bioequivalence](https://www.fda.gov/media/70958/download):

> Sample sizes for average BE should be obtained using published formulas. Sample sizes for population and individual BE should be based on simulated data. The simulations should be conducted using a default situation allowing the two formulations to vary as much as 5% in average BA with equal variances and certain magnitude of subject-by-formulation interaction. The study should have 80 or 90% power to conclude BE between these two formulations. Sample size also depends on the magnitude of variability and the design of the study. Variance estimates to determine the number of subjects for a specific drug can be obtained from the biomedical literature and/or pilot studies.

Appropriate method is described in `Diletti D, Hauschke D, Steinijans VW. Sample Size Determination for Bioequivalence Assessment by Means of Confidence Intervals. Int J Clin Pharmacol Ther Toxicol. 1991;29(1):1–8` and implemented in R package [PowerTOST](https://cran.r-project.org/web/packages/PowerTOST/index.html) with one clarification: it is simulation-based (iterative) procedure rather than simple calculation by formula.

```{r}
library(PowerTOST)
library(knitr)
library(data.table)
library(purrr)
```


## Sample size for standard crossover design (2x2x2) and 4 period full replicate design (2x2x4)

`sampleN.TOST()` function can calculate sample size for different designs:

```{r}
kable(known.designs())
```

Basic usage: we should specify `targetpower` (power to achieve at least, e.g. 0.8 or 0.9), `theta0` (T/R ratio if `logscale = TRUE` which is convenient default value) and `cv` (coefficient of variation given as ratio if `logscale = TRUE`).

```{r}
# 2x2x2
PowerTOST::sampleN.TOST(
  targetpower = 0.8,
  theta0 = 0.95,
  CV = 0.3,
  design = "2x2x2"
)
```

```{r}
# 2x2x4
PowerTOST::sampleN.TOST(
  targetpower = 0.9,
  theta0 = 0.98,
  CV = 0.24,
  design = "2x2x4"
)
```

Note that total (not per-sequence) sample size is given.

`alpha` (one-sided significance level, default is 0.05) almost never needs to be changed, `theta1` (lower bioequivalence limit) and `theta2` (upper bioequivalence limit) can be changed for non-standard bioequivalence limits, e.g. for narrow therapeutic index drugs.

## Reproduction of Table 1 from FDA Guidance for Industry. Statistical Approaches to Establishing Bioequivalence

Reproduction of Table 1 from [FDA Guidance for Industry. Statistical Approaches to Establishing Bioequivalence](https://www.fda.gov/media/70958/download) is quite tricky because it consists one more parameter to consider - the subject-by-formulation interaction variance component, $\sigma_D^2$. 

$$\sigma_D^2=(\sigma_{BT}-\sigma_{BR})^2+2\times(1-\rho)\times\sigma_{BT}\times\sigma_{BR}$$
where $\sigma_{BT}^2$ and $\sigma_{BR}^2$ are between-subject variances for the T and R formulations, respectively and $\rho$ is correlation between subject-specific means $\mu_{Tj}$ and $\mu_{Rj}$. These parameters are rarely reported in publications and can't be estimated from CI boundaries and sample size. In such lack of information one can assume $\sigma_{BT}=\sigma_{BR}$ as well as $\rho=1$. Under these reasonable assumptions $\sigma_D^2=\sigma_D=0$, so `sampleN.TOST()` calculation should be correct. 

```{r}
targetpower <- c(0.8, 0.9)
theta0 <- 1 - 0.05
CV <- c(0.15, 0.23, 0.3, 0.5)
design <- c("2x2x2", "2x2x4")

dt <- data.table::CJ(CV, targetpower, design, theta0)

sample_size <- purrr::pmap(dt, PowerTOST::sampleN.TOST, print = FALSE)
kable(rbindlist(sample_size))
```

As we can see, calculated values are equal to the reference ones for smallest $\sigma_D=0.01$ if CV=0.15 and CV=0.23. If CV=0.30 and power 80%, sample sizes are also equal, but for other parameters combinations sample sizes are underestimated. 

Conclusion: we can trust `sampleN.TOST()`; for CV less or equal 0.30 with power 80% and for CV less or equal 0.23 with power 90% it can be considered as validated against reference from FDA guidance.


## Estimate CV from CI boundaries and sample size

CV can be calculated from CI boundaries and sample size if only these values are available:

```{r}
PowerTOST::CVfromCI(lower = 0.95, upper = 1.11, n = 38)
```


::: {.callout-note collapse="true" title="Session Info"}
```{r}
sessionInfo()
```

:::
