---
title: "<R> <Tobit Regression>"
---

# Tobit regression

## Libraries

### General
```{r}
#| output: false
#| warning: false
library(dplyr)
library(gt)
library(broom)
```

### Methodology specific
```{r}
#| output: false
#| warning: false
library(emmeans)
library(censReg)
library(survival)
library(VGAM)
```


## Tobit model

Censoring occurs when data on the dependent variable is only partially known. For example, in virology, sample results could be below the lower limit of detection (eg, 100 copies/mL) and in such a case we only know that the sample result is <100 copies/mL, but we don't know the exact value.

Let $y^{*}$ be the the true underlying latent variable, and $y$ the observed variable. We discuss here censoring on the left:

$$
y =
\begin{cases}
 y^{*}, & y^{*} > \tau  \\
 \tau, & y^{*} \leq \tau
 \end{cases}       
$$
We consider tobit regression with a censored normal distribution. The model equation is
$$
y_{i}^{*} = X_{i}\beta + \epsilon_{i} 
$$
with $\epsilon_{i} \sim N(0,\sigma^2)$. But we only observe $y = max(\tau, y^{*})$.
The tobit model uses maximum likelihood estimation (for details see for example Breen, 1996). It is important to note that $\beta$ estimates the effect of $x$ on the latent variable $y^{*}$, and not on the observed value $y$.


## Data used

We assume two equally sized groups (n=10 in each group). The data is censored on the left at a value of $\tau=8.0$.
In group A 4/10 records are censored, and 1/10 in group B.

```{r create data}
dat_used = tribble(
  ~ID, ~ARM, ~Y, ~CENS,
  "001", "A", 8.0, 1, 
  "002", "A", 8.0, 1,
  "003", "A", 8.0, 1,
  "004", "A", 8.0, 1,
  "005", "A", 8.9, 0,
  "006", "A", 9.5, 0,
  "007", "A", 9.9, 0,
  "008", "A", 10.3, 0,
  "009", "A", 11.0, 0,
  "010", "A", 11.2, 0,
  "011", "B", 8.0, 1, 
  "012", "B", 9.2, 0,
  "013", "B", 9.9, 0,
  "014", "B", 10.0, 0,
  "015", "B", 10.6, 0,
  "016", "B", 10.6, 0,
  "017", "B", 11.3, 0,
  "018", "B", 11.8, 0,
  "019", "B", 12.9, 0,
  "020", "B", 13.0, 0,
)

gt(dat_used)
```


## Example Code using R

The analysis will be based on a Tobit analysis of variance with $Y$, rounded to 1 decimal places, as dependent variable and study group as a fixed covariate. A normally distributed error term will be used. Values will be left censored at the value 8.0.

Several R functions and packages are presented.

### censReg
The `censReg` function from the `censReg` package performs tobit models for left and right censored. The model is estimated by Maximum Likelihood (ML) assuming a Gaussian (normal) distribution of the error term. The maximization of the likelihood function is done by function `maxLik` of the `maxLik` package. The optimization method can be changed.
```{r censReg}
res_censreg = censReg(Y ~ ARM, 
                      left = 8.0, 
                      data = dat_used)
summary(res_censreg)

# Difference between groups (Wald CIs)
round(res_censreg$estimate[2], 3) 
round(stats::confint(res_censreg, level = 0.95)[2,], 3)
```
The output provides an estimate of difference between groups A and B (B-A), namely 1.8225 (se=0.8061). The presented p-value is a two-sided p-value based on the Z-test. The output also provides an estimate for $log(\sigma) = 0.5491$. Wald based confidence intervals can be obtained by the `stats::confint` function.



### survreg
Using the `survreg` function from the `survival` package a tobit model can be fit. For more information, refer to the [survival package](https://cran.r-project.org/web/packages/survival/index.html).
```{r survreg}
res_survreg = survreg(Surv(Y, 1-CENS, type="left") ~ ARM,
                      dist = "gaussian",
                      data = dat_used)
summary(res_survreg)

# Least square means by group
lsm = emmeans(res_survreg, specs = trt.vs.ctrl ~ ARM)
lsm$emmeans

# Difference between groups
lsm_contrast = broom::tidy(lsm$contrasts, conf.int=TRUE, conf.level=0.95)
gt(lsm_contrast) %>%
  fmt_number(decimals = 3)

# Wald-based CIs
round(stats::confint(res_survreg, level = 0.95)[2,], 3)
```

The output provides an estimate of difference between groups A and B (B-A), namely 1.823 (se=0.806). The presented p-value is a two-sided p-value based on the Z-test. The output also provides an estimate for $log(\sigma) = 0.549$. Using the `emmeans` package/function least square means and contrast can be easily obtained. The confidence intervals and p-value is based on the t-test using `emmeans`. Wald based confidence intervals can be obtained by the `stats::confint` function.


### vglm
The `VGAM` package provides functions for fitting vector generalized linear and additive models (VGLMs and VGAMs). This package centers on the iteratively reweighted least squares (IRLS) algorithm. The `vglm` function offers the possibility to fit a tobit model.
```{r vglm}
res_vglm = vglm(Y ~ ARM,
                tobit(Lower = 8.0),
                data = dat_used)
summary(res_vglm)

# Difference between groups
round(res_vglm@coefficients[3], 3)
round(confintvglm(res_vglm, level = 0.95)[3,], 3)
```

The output provides an estimate of difference between groups A and B (B-A), namely 1.823 (se=0.794). The presented p-value is a two-sided p-value based on the Z-test. Note that point estimate for the difference (and associated SE) are slightly different from the results obtained by `censReg` and `tobit` due to the difference in estimation procedure used. Wald based confidence intervals can be obtained by the `confintvglm` function. The $(Intercept):2$ in the model output is an estimate for $log(\sigma)$.

## Discussion
The results from the `censReg::censReg` and `survival::survreg` are similar. The `survival::survreg` allows for easy incorporation with the `emmeans` package (note: be aware that the standard approach with emmeans is based on the t-test and not the Z-test).

The `VGAM::vglm` approach provides slightly different results. This difference comes from the fact that a iteratively reweighted least squares (IRLS) algorithm is used for estimation.


## Reference

Breen, R. (1996). Regression models. SAGE Publications, Inc., https://doi.org/10.4135/9781412985611

Tobin, James (1958). "Estimation of Relationships for Limited Dependent Variables". Econometrica. 26 (1): 24-36. doi:10.2307/1907382




::: {.callout-note collapse="true" title="Session Info"}
```{r}
#| echo: false

# List all the packages needed 
si <- sessioninfo::session_info(c(
  #Create a vector of all the packages used in this file 
))
si
```
:::
