---
title: "Wilcoxon signed-rank test"
---

### **Introduction**

Wilcoxon signed-rank test is a non-parametric test which is sometimes used instead of the paired Student's t-test when assumptions regarding a normal distribution are not valid. It is a rank test, designed for analyzing repeated measures or paired observations by a paired comparison (a type of location test) to assess whether their population means differ. Whilst it does not 'compare' means or medians for a set of paired data, it ranks the results on A and ranks the results on B, then compares if Prob(A\>B) \> Prob(B\>A).

*Ties* are when you have two observations with the same result. For example, in a 2-period cross-over study, you take the difference between result on Treatment A minus result on Treatment B and find that two or more subjects have the same difference.

Additionally, *"0s"* can cause some trouble as well. For example when the difference between result on Treatment A minus result on Treatment B equals 0.

```{r}
#| echo: false
#| include: false
library(tidyverse)
library(broom)
library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stats)
library(coin)
library(DOS2)
library(MASS)

knitr::opts_chunk$set(echo = TRUE)

blood_p <- read.csv("../data/blood_pressure.csv", dec = ".")[1:240,1:5]

```

### Data {.unnumbered}

Analysis will be conducted on the example of anonymized data from 2-period, cross-over study comparing treatments A and B in patients with asthma and acute airway obstruction induced by repeated mannitol challenges.

Wilcoxon signed rank test was applied to analyse the time to return to baseline FEV1 post-mannitol challenge 2. Median difference, p value and 95% CI were provided using the Hodges-Lehmann estimate.

```{r}
head(blood_p)
```

### Dataset without ties {.unnumbered}

Let's consider a case where the dataset has no ties.

### Available packages {.unnumbered}

In R Wilcoxon signed rank test can be performed using for example **DOS (version 0.5.2)** or **stats (version 3.6.2)** package.

#### stats {.unnumbered}

Function *wilcox.test* used for Wilcoxon Rank Sum and Signed Rank Tests will be applied. For more information about that function go [here](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/wilcox.test)

We will focus on the below arguments: - alternative - paired - exact - correct - conf.int.

### Examples {.unnumbered}

```{r}
# Exact 
stats::wilcox.test(x = blood_p$bp_after, y = blood_p$bp_before, 
                                   paired = TRUE, 
                                   conf.int = TRUE, 
                                   conf.level = 0.9, 
                                   alterative = "two.sided", 
                                   exact = TRUE)

# No exact & continuity correction
stats::wilcox.test(x = blood_p$bp_after, y = blood_p$bp_before, 
                                   paired = TRUE, 
                                   conf.int = TRUE, 
                                   conf.level = 0.9, 
                                   alterative = "two.sided", 
                                   exact = FALSE, 
                                   correct = TRUE)


# No exact & No continuity correction
stats::wilcox.test(x = blood_p$bp_after, y = blood_p$bp_before, 
                                     paired = TRUE, 
                                     conf.int = TRUE, 
                                     conf.level = 0.9, 
                                     alterative = "two.sided" , 
                                     exact = FALSE, 
                                     correct = FALSE)

```

### Important notes on stats:wilcox.test {.unnumbered}

-   By default an exact p-value is computed if the samples size is less than 50 and there are no ties. Otherwise, a normal approximation is used.
-   If exact p-values are available, an exact confidence interval is obtained by the algorithm described in Bauer (1972), and the Hodges-Lehmann estimator is employed. Otherwise, the returned confidence interval and point estimate are based on normal approximations.
-   If non-exact p-value is calculated, continuity correction in the normal approximation for the p-value can be applied with *correct* argument.
-   Statistic V is provided, which is a test statistic based on Sprent (1993) algorithm

#### DOS2 {.unnumbered}

Function *senWilcox* used for Sensitivity Analysis for Wilcoxon's Signed-rank Statistic will be applied. For more information about that function go [here](https://www.rdocumentation.org/packages/DOS/versions/1.0.0/topics/senWilcox)

### Examples {.unnumbered}

```{r}
DOS2::senWilcox(blood_p$bp_after - blood_p$bp_before, 
                   gamma = 1, 
                   conf.int = TRUE, 
                   alpha = 0.1, 
                   alternative = "twosided")
```

### Important notes on DOS2:senWilcox {.unnumbered}

-   Gamma \>= 1 is the value of the sensitivity parameter. If gamma=1, then you are assuming ignorable treatment assignment or equivalently no unmeasured confounding - that is the considered scenario in our example, sensitivity analysis is not performed.
-   Only p value, estimate and CI are provided

### Coin package - coming soon! {.unnumbered}
